---
title: "CDC BRFSS Analysis: Veteran Health"

author: "Elsa Li, Amanda Louie, Anika Sharma"
format:
  pdf:
    geometry:
      - top=20mm
      - right=20mm
      - left=20mm
      - bottom=20mm
    knitr:
      opts_chunk: 
        out-width: "90%"
        fig-align: "center"
        size: "scriptsize"
    fontsize: 10pt
    toc: false
    toc-params: "allowframebreaks"
    include-in-header:
      - text: |
          \usepackage{etoolbox}
          \usepackage{hyperref}
          \usepackage{booktabs}
          \usepackage{array}
          \usepackage{makecell}
          \usepackage{longtable}
          \usepackage{float}
          
execute:
  echo: false
  warning: false
  message: false
  eval: true

fig-align: center
fig-height: 6
fig-width: 8
---
```{r}
#| echo: false
library(dplyr)
brfss = read.csv("brfss2022_m62.csv", stringsAsFactors=TRUE)
veteran <- subset(brfss, veteran == "Yes")
nonVeteran <- subset(brfss, veteran == "No")
```

# Introduction

Veterans make up about 15.8 million people, as of 2023, as part of the US population (US Census Bureau 2022). Health policies surrounding veterans are often a political issue that needs addressing as veterans serving experience impacts their health and a lack of support and access to their health benefits when they come back from their service affects their physical and mental wellbeing. According to National Veterans Homeless Support, “During the period from 2001 to 2020, the frequency of mental health issues and substance use disorders among Veterans Health Administration (VHA) participants increased significantly, from 27.9% to 41.9%. Veterans face a heightened risk of suicide, being 1.5 times more likely than civilians” (NVHS). We were particularly interested in investigating and characterizing the health of veterans in depth. Hence, this report is centered around the broader issue of veteran well-being. We will use the BRFSS 2022 dataset, which is a survey of risk assessment factors conducted by the CDC in the general US population (CDC 2022), to investigate veteran well-being in detail. The key categories that we will be exploring are:

1. How do the demographics of the veteran population differ from the national distribution?
2. How is veteran status associated with individuals’ physical health and mental health?
3. What are some factors that influence both mental and physical health amongst veterans?

In answering these questions and henceforth further unpacking the current state of veteran well-being, we hope we can make concrete and statistically backed claims to guide policy which better supports veterans in access, equity, and welfare. Note that all the claims and analysis we make are based on this BRFSS dataset only, and should not be extrapolated beyond without further statistical analysis.


# Demographics

When thinking about veterans in the larger context of this dataset, we were first interested in exploring *who* the veterans in the United States are. This question is important because in order to support veteran well-being we need to first understand what their demographic background is. The first step we took was to conduct exploratory data analysis on the veteran sample within the BRFSS dataset. Some notable findings we found were that 88.6% of veterans are men, the majority race is White-only, non Hispanic — making up 81.3% of the sample, 43.1% of veterans are college graduates, and 17.2% of veterans are 80 or older.

These results motivated the key guiding question: how do the demographics of veterans differ from the general population? To investigate this, we conducted a Chi-squared goodness of fit test which compared the distributions of the demographics of veterans to that of the national distribution. By demographics, here we chose to include: race, education level, and income level — we conducted separate tests for each one of these. The national average data was taken from census bureau data collected in 2022 (USAFacts 2026). The null hypothesis for the Chi-squared goodness of fit test is that the veteran sample follows the same distribution as the national average and the alternative hypothesis is that the veteran sample does not follow the same distribution as the national average.

Let’s consider the assumptions of a Chi-squared test. First, we assume that there is independence between outcomes. This means that data points aren’t paired, and are independent of one another. Since the BRFSS dataset comes from a large sampled survey, we can assume random sampling where each respondent’s answers is independent of others. The second assumption is that the sample size must have at least 5 expected cases. This is true for all the demographic variables since we have $n = 42841$, so the expected counts $(n \times$ expected proportion). is always larger than 5.  The last condition is that $df  = k-1 > 1$ (where $k$ is the number of categories) which holds since we have more than 2 categories for each of our demographic variables. Hence, all the conditions are met.

After conducting a Chi-squared test, we found that in each case (race, education, and income) the results were significant at a 95% confidence level — meaning that the p-value was below 0.05.  Hence, we investigated further to see which groups differed the most using multiple 1-sample proportion tests with adjusted p-values using the Bonferroni correction. This test (along with the same independence assumption stated above) assumes that $Z \sim N(0,1)$ meaning that the Z-distribution can be approximated by the normal distribution. This is true when $np_0 \geq 10$ and $n(1-p_0) \geq 10$. Since  $n = 42841$, this condition holds for all our values of $p_0$ in each of race, education level, and income level. The results for each of the tests is shown below:

```{=latex}
\begin{table}[h]
\centering
\renewcommand{\arraystretch}{0.8}
\begin{tabular}{lrrrl}
\toprule
Race & \makecell{Veteran\\Prop.} & \makecell{National\\Prop.} & \makecell{p-value\\(adj)} & Direction \\
\midrule
Asian only, non-Hispanic & 0.0127 & 0.0610 & $< 2.2\mathrm{e}{-16}$ & Underrepresented \\
Hispanic & 0.0500 & 0.1910 & $< 2.2\mathrm{e}{-16}$ & Underrepresented \\
White only, non-Hispanic & 0.8132 & 0.5890 & $< 2.2\mathrm{e}{-16}$ & Overrepresented \\
Black only, non-Hispanic & 0.0743 & 0.1260 & $< 2.2\mathrm{e}{-16}$ & Underrepresented \\
American Indian or Alaskan Native only, Non-Hispanic & 0.0171 & 0.0070 & $< 2.2\mathrm{e}{-16}$ & Overrepresented \\
Native Hawaiian or other Pacific Islander only, Non-Hispanic & 0.0063 & 0.0020 & $< 2.2\mathrm{e}{-16}$ & Overrepresented \\
Multiracial, non-Hispanic & 0.0264 & 0.0240 & $1.16 \times 10^{-2}$ & Overrepresented \\
\bottomrule
\end{tabular}
\end{table}

\begin{center}

\textbf{Figure 1:} 1-sample proportion test between the national average and the sampled veteran population (race)

\end{center}
```


```{=latex}
\begin{table}[h]
\centering
\renewcommand{\arraystretch}{0.8}
\begin{tabular}{lrrrl}
\toprule
Education & Observed\_Prop & Expected\_Prop & p\_bonferroni & Direction \\
\midrule
Less than high school & 0.02516292 & 0.1088911 & 0.000000e+00 & Underrepresented \\
Bachelor's or higher & 0.42279884 & 0.3426573 & 4.203017e-262 & Overrepresented \\
Some college / associate & 0.32483470 & 0.2847153 & 1.206677e-73 & Overrepresented \\
High school graduate & 0.22720354 & 0.2637363 & 3.298299e-64 & Underrepresented \\
\bottomrule
\end{tabular}
\end{table}

\begin{center}

\textbf{Figure 2:} 1-sample proportion test between the national average and the sampled veteran population (education)

\end{center}
```

```{=latex}
\begin{table}[h]
\centering
\renewcommand{\arraystretch}{0.8}
\begin{tabular}{lrrrl}
\toprule
Income & Observed\_Prop & Expected\_Prop & p\_bonferroni & Direction \\
\midrule
\$25,000 to less than \$50,000 & 0.2661385 & 0.171 & 0.000000e+00 & Overrepresented \\
\$75,000 or more & 0.4214797 & 0.520 & 7.461731e-310 & Underrepresented \\
\$50,000 to less than \$75,000 & 0.2022149 & 0.148 & 2.430662e-186 & Overrepresented \\
Less than \$25,000 & 0.1101669 & 0.161 & 3.483507e-153 & Underrepresented \\
\bottomrule
\end{tabular}
\end{table}

\begin{center}

\textbf{Figure 3:} 1-sample proportion test between the national average and the sampled veteran population (income level).

\end{center}
```

In all of these cases, we can see that the adjusted p-values are still well under 0.05. Hence this suggests that there is statistically significant evidence that the population proportion is different from the expected value for each group. We also compute the direction of the difference: some groups are underrepresented whereas others are overrepresented. Through this we see some expected trends following our exploratory analysis. For example, the group White only-non hispanic is overrepresented. Although we also see some perhaps counterintuitive trends. For example, the group of Bachelor’s or higher education level is overrepresented, but the group earning $75k or more is underrepresented. It’s important to note also that statistical significance doesn't always translate to practical significance. An example of this is in the race demographic where the value we observed for Multiracial, non-Hispanic was 0.026, and what we expected was 0.024 — which practically may not be significant. Also, since the sampled veteran population is skewed in terms of some demographic factors (Eg: veterans in this dataset are much older than the general population), the comparison to the entire US population data should be treated with caution.

Overall, we can conclude that the demographics of the veteran population differ from the national population, and it motivates the next question: how might these demographic differences affect comparisons of health outcomes between veterans and non-veterans? 

# Physical and Mental Health in Veterans vs. Non-Veterans

We looked how being a veteran affects the likelihood of being susceptible to certain physical and mental health factors. The significance of this test determines whether we believe health policies should vary between veterans and non-veterans. In order to do this, we ran proportion comparison z-tests and a two-sample mean comparison t-test between the veterans and non-veterans that experience certain physical and mental health factors. The results of our comparison tests are discussed in the following sections.

## Physical Health Comparisons

In our comparison tests, which were two-sample proportion z-tests and a two-sample mean comparison test, our proposed null hypothesis is that the proportion of veterans that experience specific physical health factors was not different from the proportion of non-veterans who experience the same factors ($H_0: p_{veteran} = p_{nonveteran}$, $H_A: p_{veteran} \neq p_{nonveteran}$).For the category `daysphyshlthnotgood`, participants reported the number of days in the past month where their physical health was not good and so we compared the means in this case instead.
 
The assumptions we made while conducting our comparison tests were that the veterans and non-veterans datasets are independent and individual people are independent from each other. These conditions are both satisfied since being a veteran is independent of being a non-veteran and each participant in the dataset answered the survey independent of each other. Another assumption that our proportion comparison tests made was that the success–failure condition, $np ≥ 10$ and $n(1-p) ≥ 10$, was satisfied, where n is the size of the sample and p is the proportion. For this dataset and proportions the success-failure condition was satisfied since for all of the proportions used in this test multiplying the proportion and 1 minus the proportion by the sample size, as determined in section 1, would result in a number larger than 10. Two additional assumptions our mean comparison test made were that the sample means were normally distributed among veterans and non-veterans and that the datasets had equal variances. The first assumption is satisfied by the Central Limit Theorem and the sizes of each dataset as defined in section 1, leading us to conclude that the sample means are normally distributed. We checked the second assumption by plotting a histogram of the “daysphyshlthnotgood” for veteran and non-veterans and got the following plots with similar spread of data, which led us to believe that the variances were equal. In our initial testing, we conducted our comparison tests with equal and unequal variances and found that we came to the same conclusion about the significance of any difference in the means through both tests.

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 4

veteranDaysphyshlthnotgoodClean <- veteran %>% filter(!is.na(daysphyshlthnotgood))
nonVeteranDaysphyshlthnotgoodClean <- nonVeteran %>% filter(!is.na(daysphyshlthnotgood))

par(mfrow = c(1,2))

hist(veteranDaysphyshlthnotgoodClean$daysphyshlthnotgood,
     main = "Days Physical Health Not Good (Veteran)",
     xlab = "Days in past month phys. health not good")
hist(nonVeteranDaysphyshlthnotgoodClean$daysphyshlthnotgood,
     main = "Days Physical Health Not Good (Non-Veterans)",
     xlab = "Days in past month phys. health not good")
```

```{=latex}
\begin{center}
\textbf{Figure 4:} Histograms of the distribution of `daysphyshlthnotgood` in veterans and non-veterans to visualize variance among datasets
\end{center}
```

The results of our t-tests are summarized in the table below.

```{=latex}
\begin{table}[h]
\centering
\renewcommand{\arraystretch}{0.8}
\begin{tabular}{lccc}
\toprule
Factor & Veteran & Non-Veteran & $p$-value \\
\midrule
Arthritis & 0.4356422 & 0.3455287 & $< 2.2 \times 10^{-16}$ \\
Heart Disease & 0.1244373 & 0.0543102 & $< 2.2 \times 10^{-16}$ \\
Cancer & 0.1789827 & 0.1105903 & $< 2.2 \times 10^{-16}$ \\
Diabetes & 0.1962027 & 0.1340290 & $< 2.2 \times 10^{-16}$ \\
COPD & 0.11969946 & 0.07724301 & $< 2.2 \times 10^{-16}$ \\
Days Physical Health Not Good & 4.972634 & 4.338173 & $< 2.2 \times 10^{-16}$ \\
\bottomrule
\end{tabular}
\end{table}

\begin{center}
\textbf{Figure 5:} Results of comparison tests between veterans and non-veterans on different physical health factors
\end{center}
```

From these comparison tests, we can conclude that there is a significant difference in the proportions (and mean) of these physical health factors present in veterans versus non-veterans. All of our comparison tests resulted in a p-value less than $2.2 \times 10^-16$, which indicates that we would reject our null hypothesis that the proportions or means are equal. We notice that in general the proportion of veterans who are diagnosed with these physical health factors is greater than the proportion of non-veterans diagnosed with the same physical health factors. We can visualize this difference in this barplot that compares the different proportions among veterans and non-veterans.


```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 4
# generated with help from gemini

# --- Data ---
labels_a <- c("Arthritis", "Heart Disease", "Cancer", "Diabetes", "COPD")
prop_vet <- c(0.4356422, 0.1244373, 0.1789827, 0.1962027, 0.11969946)
prop_nonVet <- c(0.3455287, 0.0543102, 0.1105903, 0.1340290, 0.07724301)
data_matrix_a <- rbind(prop_vet, prop_nonVet)
colnames(data_matrix_a) <- labels_a

labels_b <- c("Days Phys. Health\nNot Good")
mean_vet <- c(4.972634)
mean_nonVet <- c(4.338173)
data_matrix_b <- rbind(mean_vet, mean_nonVet)
colnames(data_matrix_b) <- labels_b

chart_colors <- c("#2a79d5", "#29b9d6")

# --- Layout Configuration ---
par(mfrow = c(1, 2), mar = c(5, 4, 4, 2))

# --- Plot 1: Chronic Conditions ---
# Initialize the plot area without drawing bars yet to set up the grid
bp1 <- barplot(data_matrix_a, beside = TRUE, col = NA, border = NA, 
               main = "Physical Health", ylab = "Proportion", 
               ylim = c(0, 0.5), las = 1, cex.names = 0.7, font.main = 2)

# Draw the grid lines FIRST
grid(nx = NA, ny = NULL, col = "gray90", lty = "solid")

# Overlay the bars ON TOP of the grid
barplot(data_matrix_a, 
        beside = TRUE, 
        col = chart_colors,
        border = "white",
        add = TRUE,          # This tells R to draw over the existing grid
        las = 1, 
        cex.names = 0.7,
        xaxt = "n")          # Prevents double-drawing labels

legend("topright", legend = c("Veteran", "Non-Veteran"), 
       fill = chart_colors, bty = "n", cex = 0.7)

# --- Plot 2: Quality of Life ---
barplot(data_matrix_b, beside = TRUE, col = NA, border = NA, 
        main = "Average Days Unwell", ylab = "Mean Days", 
        ylim = c(0, 6), las = 1, cex.names = 0.7, font.main = 2)

grid(nx = NA, ny = NULL, col = "gray90", lty = "solid")

barplot(data_matrix_b, 
        beside = TRUE, 
        col = chart_colors,
        border = "white",
        add = TRUE, 
        las = 2, 
        cex.names = 0.7,
        xaxt = "n")

# Reset
par(mfrow = c(1, 1))
```

```{=latex}
\textbf{Figure 6:} Differences in proportions between veterans and non-veterans on different physical health factors
```

## Mental Health Comparisons

We proposed was the same comparison tests for mental health as for physical health factors $(H_0: p_{veteran} = p_{nonveteran}, H_A: p_{veteran} \neq p_{nonveteran})$. The assumptions made in the proportion and mean comparison test were the same and verified the same way. To verify that the datasets had equal variances we plotted a histogram of the `daysmenthlthnotgood` for veteran and non-veterans and got plots with similar spread of data, which led us to believe that the variances were equal. In our initial testing, we conducted our two-sample t-tests with equal and unequal variances and found that we came to the same conclusion about the significance of any difference in the means through both tests.

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 4

veteranDaysmenthlthnotgoodClean <- veteran %>% filter(!is.na(daysmenthlthnotgood))
nonVeteranDaysmenthlthnotgoodClean <- nonVeteran %>% filter(!is.na(daysmenthlthnotgood))

par(mfrow = c(1,2))

hist(veteranDaysmenthlthnotgoodClean$daysmenthlthnotgood,
     main = "Days Mental Health Not Good (Veteran)",
     xlab = "Number of days in past month mental health was not good")
hist(nonVeteranDaysmenthlthnotgoodClean$daysmenthlthnotgood,
     main = "Days Mental Health Not Good (Non-Veteran)",
     xlab = "Number of days in past month mental health was not good")
```

```{=latex}
\begin{center}
\textbf{Figure 7:} Histograms of the distribution of “daysmenthlthnotgood” in veterans and non-veterans to visualize variance among datasets
\end{center}
```

The results of our comparison tests are summarized in the table below.

```{=latex}
\begin{table}[h]
\centering
\renewcommand{\arraystretch}{0.8}
\begin{tabular}{lccc}
\toprule
Factor & Veteran & Non-Veteran & $p$-value \\
\midrule
Depression & 0.1757344 & 0.2166535 & $< 2.2 \times 10^{-16}$ \\
Days Mental Health Not Good & 3.517690 & 4.462462 & $< 2.2 \times 10^{-16}$ \\
\bottomrule
\end{tabular}
\end{table}

\begin{center}
\textbf{Figure 8:} Differences in proportions between veterans and non-veterans on different mental health factors
\end{center}
```

Similarly to physical health, from these tests we can conclude that there is a significant difference in the proportion of veterans and the proportion of non-veterans who have been diagnosed with a depressive disorder. Each of our comparison tests resulted in a p-value less than $2.2 \times 10^-16$, which indicates that we would reject our null hypothesis that the proportions or means are the same. We can notice that in general, unlike with the physical health factors, the proportion of veterans who are diagnosed with a depressive disorder is less than the proportion of non-veterans diagnosed with a depressive disorder. We can visualize this difference in this barplot that compares the different proportions among veterans and non-veterans.

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 4
# generated with help from gemini

labels_a <- c("Depression")
prop_vet <- c(0.1757344)
prop_nonVet <- c(0.2166535)
data_matrix_a <- rbind(prop_vet, prop_nonVet)
colnames(data_matrix_a) <- labels_a

labels_b <- c("Days Mental Health\nNot Good")
mean_vet <- c(3.517690)
mean_nonVet <- c(4.462462)
data_matrix_b <- rbind(mean_vet, mean_nonVet)
colnames(data_matrix_b) <- labels_b

chart_colors <- c("#2a79d5", "#29b9d6")

# --- Layout Configuration ---
par(mfrow = c(1, 2), mar = c(5, 4, 4, 2))

# --- Plot 1: Chronic Conditions ---
# Initialize the plot area without drawing bars yet to set up the grid
bp1 <- barplot(data_matrix_a, beside = TRUE, col = NA, border = NA, 
               main = "Depression", ylab = "Proportion", 
               ylim = c(0, 0.5), las = 1, cex.names = 0.7, font.main = 2)

# Draw the grid lines FIRST
grid(nx = NA, ny = NULL, col = "gray90", lty = "solid")

# Overlay the bars ON TOP of the grid
barplot(data_matrix_a, 
        beside = TRUE, 
        col = chart_colors,
        border = "white",
        add = TRUE,          # This tells R to draw over the existing grid
        las = 1, 
        cex.names = 0.7,
        xaxt = "n")          # Prevents double-drawing labels

legend("topright", legend = c("Veteran", "Non-Veteran"), 
       fill = chart_colors, bty = "n", cex = 0.7)

# --- Plot 2: Quality of Life ---
barplot(data_matrix_b, beside = TRUE, col = NA, border = NA, 
        main = "Average Days Unwell", ylab = "Mean Days", 
        ylim = c(0, 6), las = 1, cex.names = 0.7, font.main = 2)

grid(nx = NA, ny = NULL, col = "gray90", lty = "solid")

barplot(data_matrix_b, 
        beside = TRUE, 
        col = chart_colors,
        border = "white",
        add = TRUE, 
        las = 2, 
        cex.names = 0.7,
        xaxt = "n")

# Reset
par(mfrow = c(1, 1))
```

```{=latex}
\begin{center}
\textbf{Figure 9:} Differences in proportions between veterans and non-veterans on different mental health factors
\end{center}
```

# Factors Affecting Physical and Mental Health of Veterans

Within the veteran population, we were interested in finding out what makes veterans prone to poor mental and physical health, which would inform measures to support veteran wellbeing effectively. For the physical health of veterans, we were most interested in finding the factors that significantly predict the likelihood of cancer, heart disease, diabetes, and BMI. Cancer, heart disease, and diabetes are all binary variables, so we modeled them using logistic regression. Since BMI is a numerical variable, we modelled it using a multiple linear regression model. The null hypothesis for both the logistic regressions and the linear regression are the same: for each predictor $x_i$, the corresponding $\beta_i = 0$, where $\beta_i$ is the coefficient to $x_i$,  in the regression model. The alternative hypothesis for the regression models is also the same: There is at least one predictor $x_i$, for which the corresponding $\beta_i \neq 0$, where $\beta_i$ is the coefficient to $x_i$, in the regression model. 

To narrow down predictors for each model, we choose a set of predictors that we believed would most significantly attribute to whatever the model was trying to predict. Although we could have used the stepwise function to eliminate those predictors that were not significant for predicting depressive disorder, we decided against doing so. This is because p-values and confidence intervals may no longer be valid after running the stepwise selection function, and we wanted to interpret the model using these features of the output. Additionally, we modified the age column to consolidate age into either young adult, middle age, and senior, and the income columns into either low, middle, or high income. For heart disease, cancer, and diabetes, our chosen set of predictors was `sex`, `bmi`, `income`, `daysphyshlthnotgood`, `anyexer`, `race`, `age`, `bingedrinker`, and `weeklydrinks`. For BMI, our set was the same just without the BMI predictor. 

We can begin with the results for our physical health models, starting with heart disease. Note that we are only displaying the predictors that had a significant adjusted p-value based on the Bonferroni correction. Since we’re using so many predictors, we decided to use this higher confidence level. One interesting result we observe here is that, keeping all other predictors constant, the odds of a male veteran having heart disease is about 2.23 times higher to that of a female veteran. 

```{=latex}
\begin{table}[h]
\centering
\renewcommand{\arraystretch}{0.8}
\begin{tabular}{l r r r}
\toprule
\textbf{Heart Disease Predictor} & Coefficient & p\_value & Odds\_Ratio \\
\midrule
(Intercept) & -4.38013938 & 1.030130e-94 & 0.01252361 \\
sexMale & 0.80261851 & 3.610160e-20 & 2.23137617 \\
bmi & 0.02059709 & 3.387576e-10 & 1.02081067 \\
incomelow income & 0.25661775 & 5.575626e-04 & 1.29255095 \\
daysphyshlthnotgood & 0.03581590 & 9.450336e-106 & 1.03646502 \\
anyexerYes & -0.18493519 & 5.283247e-06 & 0.83115815 \\
raceNative Hawaiian or other Pacific Islander only, Non-Hispanic & 0.68088540 & 7.775356e-03 & 1.97562618 \\
agesenior & 1.23859720 & 1.724281e-138 & 3.45076934 \\
ageyoung adult & -2.24758619 & 1.018028e-19 & 0.10565395 \\
bingedrinkerYes & -0.22273369 & 7.370868e-03 & 0.80032796 \\
weeklydrinks & -0.01068126 & 8.289779e-04 & 0.98937558 \\
\bottomrule
\end{tabular}
\end{table}
```

For cancer, we note that some interesting results are that being a senior was again an especially important  for predicting cancer, as it raised the odds of a senior having cancer was about 3.54 times more likely than the reference group of middle aged, when keeping all other predictors constant. 

```{=latex}
\begin{table}[h]
\centering
\renewcommand{\arraystretch}{0.8}
\begin{tabular}{l r r r}
\toprule
\textbf{Cancer Predictor} & Coefficient & p\_value & Odds\_Ratio \\
\midrule
(Intercept) & -2.03178780 & 2.882294e-30 & 0.1311009 \\
bmi & -0.01243622 & 3.430060e-05 & 0.9876408 \\
daysphyshlthnotgood & 0.02221518 & 1.349809e-46 & 1.0224638 \\
raceAsian only, non-Hispanic & -0.94885042 & 1.690569e-04 & 0.3871859 \\
agesenior & 1.31578327 & 4.803320e-215 & 3.7276696 \\
ageyoung adult & -1.28756612 & 5.411407e-24 & 0.2759416 \\
bingedrinkerYes & -0.27515240 & 1.761594e-05 & 0.7594564 \\
incomelow income & -0.23173324 & 1.067433e-04 & 0.7931577 \\
\bottomrule
\end{tabular}
\end{table}
```

Once again, being a senior was a significant contributor to having diabetes. Additionally, being a Native Hawaiian or other Pacific Islander increased a participating veteran’s odds of having diabetes by about 2.06 times versus the reference race of American Indian or Alaska Native, when keeping all other predictors constant. 

```{=latex}
\begin{table}[h]
\centering
\renewcommand{\arraystretch}{0.8}
\begin{tabular}{l r r r}
\toprule
\textbf{Diabetes Predictor} & Coefficient & p\_value & Odds\_Ratio \\
\midrule
(Intercept) & -4.52351385 & 7.071627e-171 & 0.01085083 \\
sexMale & 0.50020112 & 1.150702e-17 & 1.64905289 \\
bmi & 0.08699755 & 5.699442e-218 & 1.09089400 \\
incomelow income & 0.49233993 & 2.041857e-15 & 1.63614020 \\
incomemedium income & 0.33029307 & 4.967705e-09 & 1.39137584 \\
daysphyshlthnotgood & 0.02216352 & 1.309265e-50 & 1.02241096 \\
anyexerYes & -0.37677571 & 8.668877e-29 & 0.68606993 \\
raceWhite only, non-Hispanic & -0.44906561 & 4.238654e-05 & 0.63822422 \\
agesenior & 0.75707326 & 1.324741e-98 & 2.13202719 \\
ageyoung adult & -2.33914218 & 5.979166e-58 & 0.09641031 \\
bingedrinkerYes & -0.20653020 & 2.002169e-03 & 0.81340170 \\
weeklydrinks & -0.02587227 & 7.107947e-17 & 0.97445955 \\
\bottomrule
\end{tabular}
\end{table}
```

Again, being a senior or being a Native Hawaiian or other Pacific Islander increased a participating veteran’s predicted BMI compared to the reference race of American Indian or Alaska Native, when holding all other predictors constant. 

```{=latex}
\begin{table}[h]
\centering
\renewcommand{\arraystretch}{0.8}
\begin{tabular}{l r r r}
\toprule
\textbf{BMI Predictor} & Coefficient & p\_value \\
\midrule
(Intercept) & 30.11042063 & 0.000000e+00 \\
sexMale & 0.58101745 & 5.184119e-09 \\
incomemedium income & 0.45076399 & 2.949942e-06 \\
daysphyshlthnotgood & 0.03496438 & 3.582552e-25 \\
anyexerYes & -1.35134430 & 1.732838e-74 \\
raceAsian only, non-Hispanic & -2.29296896 & 1.295361e-10 \\
agesenior & -1.91867928 & 1.211723e-165 \\
ageyoung adult & -1.11652855 & 2.464129e-24 \\
bingedrinkerYes & 0.37072697 & 3.320259e-04 \\
weeklydrinks & -0.02155940 & 6.859478e-12 \\
\bottomrule
\end{tabular}
\end{table}
```

We can conclude by checking the assumptions for the logistic regression model. First, we assume that the observations in our dataset are independent. Since this BRFSS dataset is made up of randomly chosen respondents from all over the United States, we can assume that the veterans in the dataset are all independent of each other as well. The total number of veterans in the BRFSS dataset is 42841, and the number of veterans is also greater than 32000 within each of the factors we examined, so our sample size is sufficiently large. Lastly, we also assume that the logistic odds of success are linear in the predictors. Since we were not given the statistical tools to validate this assumption, we will not verify this assumption  in this report, but note that our conclusions may have type I error where we rejected the null hypothesis despite it being true.

Next, we can check the assumptions for the multiple linear regression model. Our assumptions are that the linear model holds, the errors are independent of each other, and the sample mean error is normally distributed. From the residual plots in Fig 10, we can see that our assumption does not hold, since there seems to be a lot more residuals above the red line, and most residuals are clustered towards higher fitted values. From the Q-Q plot, we can see a distinct right tail, which also suggests that residuals are not normally distributed. However, we also know that the central limit theorem applies because our sample size is large, but we must still interpret any final output of the linear model with caution. 

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 4

# Two lines below generated by Gemini, prompt was that given these columns, drop any rows where any of these columns are NA
# Define the variables of interest
vars <- c("heartdisease", "sex", "bmi", "cancer", "copd", 
          "daysphyshlthnotgood", "anyexer", "race", 
          "age", "bingedrinker", "weeklydrinks", "heartdisease")

# Subset rows where all specified columns are non-NA
veteran_physical <- veteran[complete.cases(veteran[, vars]), ]

model_bmi <- lm(formula = bmi ~ sex + copd +
                 daysphyshlthnotgood + anyexer + race +
                 age + bingedrinker + weeklydrinks , data=veteran_physical)

par(mfrow = c(1,2))
plot(model_bmi, which = 1)
plot(model_bmi, which = 2)
```

```{=latex}
\begin{center}
\textbf{Figure 10} Residuals Plots of BMI linear regression model 
\end{center}
```
# Factors Affecting Mental Health

For the mental health of veterans, we decided to focus on whether or not they have been told that they have a depressive disorder. Similar to how we conducted the data analysis, we ran a logistic regression model with some preselected predictors. This time, we chose our predictors to be `sex`, `income`, `race`,  `age`, `bingedrinker`,  `weeklydrinks`, `daysmenthlthnotgood`, and `anyexer`. From the results of the model, we can see that lower income veterans were around 1.9 times more likely to have been told that they have a depressive disorder than the reference group of high income veterans, holding all other predictors constant. 

```{=latex}
\begin{table}[h]
\centering
\renewcommand{\arraystretch}{0.8}
\begin{tabular}{l r r r}
\toprule
\textbf{Depression Predictor} & Coefficient & p\_value & Odds\_Ratio \\
\midrule
(Intercept) & -1.2856906 & 9.250189e-21 & 0.2764596 \\
sexMale & -0.8586733 & 6.026534e-85 & 0.4237238 \\
incomelow income & 0.6356814 & 2.922495e-23 & 1.8883083 \\
incomemedium income & 0.3306027 & 7.745822e-09 & 1.3918067 \\
daysmenthlthnotgood & 0.1072040 & 0.000000e+00 & 1.1131613 \\
anyexerYes & -0.2058627 & 5.118803e-08 & 0.8139448 \\
agesenior & -0.5226292 & 1.694792e-44 & 0.5929595 \\
\bottomrule
\end{tabular}
\end{table}
```

Lastly, we can verify our assumptions about this model. Since this is just a logistic regression model, the same assumptions about sample size and logistic odds of success being linear in the predictors hold. 

# Conclusions and Recommendations

From the results of all of the above models we gain a number of key insights. Although before we get into the results, we'd like to restate that all these conclusions should be interpreted only within the context of the BRFSS dataset and cannot be extrapolated beyond without further analysis.

From the initial demographic section we learned that the demographics of veterans are statistically different from the general US population in income, education and race. Certain races (Black, Hispanic, and Asians) were underrepresented in the veteran population, whilst others are overrepresented — such as White-only non-Hispanic. This could inform health policy and staff to be more prepared for diseases which put certain races at higher-risk than others.

In the second section, we unpacked how given a particular health factor, the difference in  proportion of veterans and non-veteran who have it is statistically significant. In particular, for physical health factors (heart disease, diabetes, cancer, BMI), we found that veterans are more likely to have the factor. Whereas, for depression, non-veterans are more likely to have it. Hence, this could inform health policy to guide more targeted attention for the physical health of veterans through prevention and holistic well-being support. In contrast, the mental health policies should remain broad and be catered to both the general population and veterans.

In the last section, we further examined the factors that impact veterans’ physical and mental health. An unsurprising result is that being a senior almost always increases a veteran’s odds of health issues the most. Additionally, we repeatedly found that being a Native Hawaiian or other Pacific Islander increased odds of having multiple health conditions. We also found that for mental health issues, lower income veterans were more likely to have been told that they have a depressive disorder. Therefore, we suggest allocating more physical health resources to the older veteran and Native Hawaiian and other Pacific Islander population, and allocate more mental health resources to lower income veteran communities. 



\newpage

## Bibliography

Bureau, US Census. “Veterans Day 2024: November 11.” Census.gov, 16 Oct. 2024, www.census.gov/newsroom/facts-for-features/2024/veterans-day.html.CDC. 

“CDC - BRFSS Annual Survey Data.” Www.cdc.gov, 31 Aug. 2020, www.cdc.gov/brfss/annual_data/annual_data.htm.USAFacts.

“How Many People Live in the US?” USAFacts, 4 Mar. 2026, usafacts.org/answers/how-many-people-live-in-the-us/country/united-states/.

“Understanding Veterans and Mental Illness.” NVHS, 16 Apr. 2026, nvhs.org/veterans-and-mental-illness/?gad_source=1&gad_campaignid=15514416182&gbraid=0AAAAADmTLu
_p7ZVXVkF4UyN7N9n5jjY9F&gclid=Cj0KCQjwkrzPBhCqARIsAJN460n3DcDuNSRaDDfYjG504YRMtHBaX3T7JY9CyThlfLlgCba47os4apEaAhUjEALw_wcB. 

\newpage

## Appendix

Below is all the code we used to complete tests and generate models. All graphs and tables come from the code from here.   



# EDA Section



```{r}
#| echo: false
#| label: fig-status-dist
#| fig-cap: "Comparison of sample sizes between Veterans and Non-Veterans."
library(ggplot2)
library(dplyr)
library(tidyverse)
library(readr)
library(dplyr)
library(ggplot2)
library(patchwork)
library(stringr)

# Define custom colors
vet_blue <- "#2a79d5"
vet_teal <- "#29b9d6"
brfss = read.csv("brfss2022_m62.csv", stringsAsFactors=TRUE)
veterans <- brfss %>% filter(veteran == "Yes")
non_veterans <- brfss %>% filter(veteran == "No")


ggplot(brfss %>% filter(!is.na(veteran)), aes(x = veteran)) +
  geom_bar(fill = vet_blue, width = 0.6) +
  geom_text(stat = "count", aes(label = scales::comma(after_stat(count))), 
            vjust = -0.5, size = 4, fontface = "bold") +
  labs(title = "Distribution of Veterans vs Non-Veterans",
       x = "Veteran Status",
       y = "Total Count") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.x = element_blank())

```



```{r}
#| echo: false
#| label: fig-log-reg
#| fig-height: 7
#| fig-width: 10

# Prepare data
model_data <- brfss %>%
  filter(!is.na(veteran)) %>%
  mutate(
    veteran_bin = as.integer(veteran == "Yes"),
    across(where(is.character), as.factor)
  ) %>%
  select(-veteran, -state, -adults) %>% 
  na.omit()

# Fit model
log_model <- glm(veteran_bin ~ ., data = model_data, family = binomial)

# Extract and format coefficients
coef_df <- as.data.frame(summary(log_model)$coefficients) %>%
  mutate(term = rownames(.)) %>%
  rename(estimate = Estimate, std_error = `Std. Error`, p_value = `Pr(>|z|)`) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    odds_ratio = exp(estimate),
    ci_low  = exp(estimate - 1.96 * std_error),
    ci_high = exp(estimate + 1.96 * std_error),
    significant = p_value < 0.05
  ) %>%
  arrange(desc(abs(estimate))) %>%
  slice_head(n = 20)

# Plot using the custom color scheme
ggplot(coef_df, aes(x = odds_ratio, y = reorder(term, odds_ratio), color = significant)) +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = ci_low, xmax = ci_high), height = 0.3) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "gray40") +
  scale_color_manual(values = c("FALSE" = "gray70", "TRUE" = vet_blue),
                     labels = c("Not significant", "p < 0.05")) +
  labs(
    title = "Top 20 Predictors of Veteran Status",
    subtitle = "Odds ratios calculated via logistic regression",
    x = "Odds Ratio (95% CI)",
    y = NULL,
    color = "Significance"
  ) +
  theme_minimal()
```


```{r}
#| echo: false
#| fig-height: 10
#| fig-width: 12
#| label: fig-vet-gender-2
#| fig-cap: "Gender distribution among surveyed veterans."

ggplot(vets, aes(x = sex)) +
  geom_bar(fill = "#29b9d6", width = 0.5) +
  geom_text(stat = "count", aes(label = scales::comma(after_stat(count))), 
            vjust = -0.5, size = 3.5, fontface = "bold") +
  labs(title = "Veteran Population by Gender", x = NULL, y = "Count") +
  theme_minimal()
```




```{r}
#| echo: false
#| label: fig-vet-age
#| fig-cap: "Age distribution of veterans, showing a heavy skew toward older demographics."

#| label: fig-vet-race
#| fig-cap: "Racial demographics of veterans within the dataset."

ggplot(vets, aes(y = reorder(race, race, function(x) length(x)))) +
  geom_bar(fill = "#2a79d5") +
  geom_text(stat = "count", aes(label = scales::comma(after_stat(count))), 
            hjust = -0.1, size = 3) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.2))) +
  labs(title = "Veteran Population by Race", x = "Count", y = NULL) +
  theme_minimal()

```

```{r}
#| echo: false
#| label: fig-education-comparison
#| fig-cap: "Education level distribution by veteran status."

brfss %>%
  filter(!is.na(veteran), !is.na(education)) %>%
  count(veteran, education) %>%
  group_by(veteran) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup() %>%
  ggplot(aes(x = education, y = prop, fill = veteran)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_text(
    aes(label = scales::percent(prop, accuracy = 0.1)),
    position = position_dodge(width = 0.7),
    vjust = -0.3,
    size = 3
  ) +
  scale_y_continuous(
    labels = scales::percent,
    expand = expansion(mult = c(0, 0.15))
  ) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 12)) +
  scale_fill_manual(values = c("No" = vet_teal, "Yes" = vet_blue)) +
  labs(
    title = "Education Distribution by Veteran Status",
    x = "Education Level",
    y = "Percent Within Group",
    fill = "Veteran"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```





```{r}
#| echo: false
#| label: fig-income-comparison
#| fig-cap: "Income level distribution by veteran status."

brfss %>%
  filter(!is.na(veteran), !is.na(income)) %>%
  count(veteran, income) %>%
  group_by(veteran) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup() %>%
  ggplot(aes(x = income, y = prop, fill = veteran)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_text(
    aes(label = scales::percent(prop, accuracy = 0.1)),
    position = position_dodge(width = 0.7),
    vjust = -0.3,
    size = 3
  ) +
  scale_y_continuous(
    labels = scales::percent,
    expand = expansion(mult = c(0, 0.15))
  ) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 12)) +
  scale_fill_manual(values = c("No" = vet_teal, "Yes" = vet_blue)) +
  labs(
    title = "Income Distribution by Veteran Status",
    x = "Income Level",
    y = "Percent Within Group",
    fill = "Veteran"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```


```{r}
#| echo: false
#| label: fig-age-comparison
#| fig-cap: "Age distribution by veteran status."

brfss %>%
  filter(!is.na(veteran), !is.na(age)) %>%
  count(veteran, age) %>%
  group_by(veteran) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup() %>%
  ggplot(aes(x = age, y = prop, fill = veteran)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_text(
    aes(label = scales::percent(prop, accuracy = 0.1)),
    position = position_dodge(width = 0.7),
    vjust = -0.3,
    size = 3
  ) +
  scale_y_continuous(
    labels = scales::percent,
    expand = expansion(mult = c(0, 0.15))
  ) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 12)) +
  scale_fill_manual(values = c("No" = vet_teal, "Yes" = vet_blue)) +
  labs(
    title = "Age Distribution by Veteran Status",
    x = "Age Group",
    y = "Percent Within Group",
    fill = "Veteran"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```


# Demographics Section


```{r}
#| echo: false
library(dplyr)
library(tidyverse)
```
```{r}

# load in BRFSS dataset:

brfss = read.csv("brfss2022_m62.csv", stringsAsFactors=TRUE)


```



```{r}
#| echo: false
# Running Chisq test to compare proportions between veterans and national data (gemini-gen)
# 1. Clean the data: Filter for Veterans and non-NA race entries
brfss_veterans <- brfss %>%
  filter(veteran == "Yes") %>%
  filter(!is.na(race))

# 2. Count the number of unique entries in the race column for veterans
vet_race_counts <- table(brfss_veterans$race)

# 3. Expected national probabilities (must sum to 1 and match table order)
expected_probs <- c(0.007, 0.061, 0.126, 0.191, 0.024, 0.002, 0.589)

# 4. Run the Chi-Squared Goodness of Fit Test
chisq.test(vet_race_counts, p = expected_probs)


```
```{r} 
# 1. Reshape the data for plotting
plot_data <- final_display %>%
  tidyr::pivot_longer(
    cols = c(Expected_Count, Observed_Count), 
    names_to = "Count_Type", 
    values_to = "Count_Value"
  )

# 2. Create the Grouped Bar Plot with Log Scale
ggplot(plot_data, aes(x = Race, y = Count_Value, fill = Count_Type)) +
  geom_bar(stat = "identity", position = "dodge") +
  # Apply the log transformation
  # 'labels = scales::comma' makes the numbers easier to read (e.g., 1,000 instead of 1e3)
#  scale_y_log10(labels = scales::comma) + 
  scale_x_discrete(labels = function(x) str_wrap(x, width = 15)) +
  theme_minimal(base_size=14) +
  labs(
    title = "Observed vs. Expected Veteran Counts by Race (Log Scale)",
    x = "Racial Group",
    y = "Count (Log 10)",
    fill = "Legend"
  ) +
  scale_fill_manual(
    values = c("Expected_Count" = "#2a79d5", "Observed_Count" = "#29b9d6"),
    labels = c("Expected", "Observed")
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor.y = element_blank() # Clean up log grid lines
  )
```


```{r}
#| echo: false
# Compare veteran race proportions against national expected proportions

observed_counts <- as.numeric(vet_race_counts)
race_names <- names(vet_race_counts)
n_total <- sum(observed_counts)

race_prop_tests <- data.frame(
  Race = race_names,
  Observed_Count = observed_counts,
  Observed_Prop = observed_counts / n_total,
  Expected_Prop = expected_probs
) %>%
  rowwise() %>%
  mutate(
    p_value = prop.test(
      x = Observed_Count,
      n = n_total,
      p = Expected_Prop,
      correct = FALSE
    )$p.value
  ) %>%
  ungroup() %>%
  mutate(
    p_adj_bonferroni = p.adjust(p_value, method = "bonferroni"),
    difference = Observed_Prop - Expected_Prop,
    abs_difference = abs(difference)
  ) %>%
  arrange(p_value)

race_prop_tests
```


```{r}
#| echo: false
race_prop_tests %>%
  mutate(direction = ifelse(difference > 0, "Overrepresented", "Underrepresented")) %>%
  select(Race, Observed_Prop, Expected_Prop, direction, p_adj_bonferroni)
```



---
title: "Veteran Demographics Analysis"
format: html
---

```{r}
#| label: setup
#| message: false
#| warning: false
#| echo: false
library(dplyr)
library(gt)

# 1. Recreating your processed dataframe based on the screenshot
# (In your actual file, you can just use 'race_prop_tests' from your previous code)
race_prop_tests <- data.frame(
  Race = c("Asian only, non-Hispanic", "Hispanic", "White only, non-Hispanic", 
           "Black only, non-Hispanic", "American Indian or Alaskan Native only, Non-Hispanic",
           "Native Hawaiian or other Pacific Islander only, Non-Hispanic", "Multiracial, non-Hispanic"),
  Observed_Prop = c(0.012736694, 0.049953993, 0.813162865, 0.074313526, 0.017119473, 0.006344133, 0.026369316),
  Expected_Prop = c(0.061, 0.191, 0.589, 0.126, 0.007, 0.002, 0.024),
  p_adj_bonferroni = c(0, 0, 0, 4.990780e-219, 1.735058e-133, 4.604660e-86, 1.158692e-02)
) %>%
  mutate(direction = ifelse(Observed_Prop > Expected_Prop, "Overrepresented", "Underrepresented"))

# 2. Generate the formatted table
race_prop_tests %>%
  gt() %>%
  # Relabel columns to match your desired style
  cols_label(
    Race = "Race",
    Observed_Prop = "Veteran Prop.",
    Expected_Prop = "National Prop.",
    direction = "Representation",
    p_adj_bonferroni = "p-value (adj)"
  ) %>%
  # Format proportions as decimals (matching your first screenshot style)
  fmt_number(
    columns = c(Observed_Prop, Expected_Prop),
    decimals = 4
  ) %>%
  # Format p-values: small values as < 2.2e-16 or scientific notation
  fmt_scientific(
    columns = p_adj_bonferroni,
    decimals = 2
  ) %>%
  # Apply the "Academic/Minimalist" styling (Thin lines top and bottom)
  tab_options(
    table.border.top.color = "black",
    table.border.top.width = px(2),
    table.border.bottom.color = "black",
    table.border.bottom.width = px(2),
    column_labels.border.bottom.color = "black",
    column_labels.border.bottom.width = px(1),
    table_body.hlines.style = "none", # Removes horizontal lines between rows
    data_row.padding = px(4)
  ) %>%
  # Align text left and numbers right
  cols_align(
    align = "left",
    columns = Race
  ) %>%
  # Handle the extremely small p-values to look like the screenshot
  text_transform(
    locations = cells_body(columns = p_adj_bonferroni, rows = p_adj_bonferroni < 2.2e-16),
    fn = function(x) "< 2.2e-16"
  )



```



```{r}
#| echo: false
# measuring differences in Demographics: education level

brfss_veterans_edu <- brfss %>%
  filter(veteran == "Yes", age >= 25, !is.na(education))

table(brfss$education, useNA = "ifany")


brfss_veterans_edu <- brfss %>%
  filter(veteran == "Yes", age >= 25, !is.na(education)) %>%
  mutate(edu_group = case_when(
    education %in% c("No school", "Elementary", "Some HS") ~ "Less than high school",
    education == "HS Grad" ~ "High school graduate",
    education == "Some College" ~ "Some college",
    education == "College Grad" ~ "Bachelor's or higher",
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(edu_group))
```


```{r}
#| echo: false
# Education comparison: BRFSS veterans vs 2022 U.S. national education distribution
## Education comparison: BRFSS veterans vs 2022 U.S. national education distribution
# Source: U.S. Census ACS 2022 S1501, population age 25+

library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)
library(readr)

# Census ACS 2022 proportions, age 25+
# These are rounded published values, so we normalize them below.
expected_edu_props <- c(
  "Less than high school" = 0.109,
  "High school graduate" = 0.264,
  "Some college / associate" = 0.285,
  "Bachelor's or higher" = 0.209 + 0.134
)

# Force probabilities to sum exactly to 1
expected_edu_props <- expected_edu_props / sum(expected_edu_props)

# Clean BRFSS veteran education data
brfss_veterans_edu <- brfss %>%
  filter(veteran == "Yes", !is.na(education)) %>%
  mutate(
    age_num = if ("age" %in% names(.)) parse_number(as.character(age)) else NA_real_,
    edu_group = case_when(
      education %in% c("No school", "Elementary", "Some HS") ~ "Less than high school",
      education == "HS Grad" ~ "High school graduate",
      education == "Some College" ~ "Some college / associate",
      education == "College Grad" ~ "Bachelor's or higher",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(edu_group)) %>%
  filter(is.na(age_num) | age_num >= 25)

# Observed counts
observed_edu_counts <- table(brfss_veterans_edu$edu_group)

# Reorder expected probabilities to match observed table
expected_edu_props_ordered <- expected_edu_props[names(observed_edu_counts)]

# Normalize again after reordering, just in case any categories are missing
expected_edu_props_ordered <- expected_edu_props_ordered / sum(expected_edu_props_ordered)

# Chi-square goodness-of-fit test
edu_chisq_test <- chisq.test(
  observed_edu_counts,
  p = expected_edu_props_ordered
)

edu_chisq_test

# Table showing which education groups differ most
edu_diff_table <- data.frame(
  Education = names(observed_edu_counts),
  Observed_Count = as.numeric(observed_edu_counts),
  Expected_Prop = expected_edu_props_ordered
) %>%
  mutate(
    Total = sum(Observed_Count),
    Observed_Prop = Observed_Count / Total,
    Expected_Count = Expected_Prop * Total,
#    Difference = Observed_Prop - Expected_Prop,
#    Percent_Point_Gap = 100 * Difference,
 #   ChiSq_Contribution = (Observed_Count - Expected_Count)^2 / Expected_Count,
    Direction = ifelse(Difference > 0, "Overrepresented", "Underrepresented")
  ) %>%
  arrange(desc(ChiSq_Contribution))

edu_diff_table

# Education group that differs most
edu_diff_table %>%
  slice_max(ChiSq_Contribution, n = 1)

# Plot observed vs expected proportions
edu_diff_table %>%
  select(Education, Observed_Prop, Expected_Prop) %>%
  pivot_longer(
    cols = c(Observed_Prop, Expected_Prop),
    names_to = "Type",
    values_to = "Proportion"
  ) %>%
  ggplot(aes(x = Education, y = Proportion, fill = Type)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Veteran Education Distribution vs 2022 U.S. National Average",
    x = "Education Level",
    y = "Proportion",
    fill = "Group"
  ) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```


```{r}
#| echo: false
# One-sample proportion tests for each education group
# with Bonferroni correction

edu_prop_tests <- data.frame(
  Education = names(observed_edu_counts),
  Observed_Count = as.numeric(observed_edu_counts),
  Expected_Prop = expected_edu_props_ordered
) %>%
  mutate(
    Total = sum(Observed_Count),
    Observed_Prop = Observed_Count / Total,
    Difference = Observed_Prop - Expected_Prop,
  ) %>%
  rowwise() %>%
  mutate(
    p_value = prop.test(
      x = Observed_Count,
      n = Total,
      p = Expected_Prop,
      correct = FALSE
    )$p.value
  ) %>%
  ungroup() %>%
  mutate(
    p_bonferroni = p.adjust(p_value, method = "bonferroni"),
    Significant_Bonferroni = p_bonferroni < 0.05,
    Direction = ifelse(Difference > 0, "Overrepresented", "Underrepresented")
  ) %>%
  arrange(p_bonferroni)

edu_prop_tests <- edu_prop_tests %>%
  select(
    Education,
    Observed_Prop,
    Expected_Prop,
    p_bonferroni,
    Direction
  )

edu_prop_tests
```


```{r}
#| echo: false
# Income comparison: BRFSS veterans vs 2022 U.S. national household income distribution

library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)

# Manually entered 2022 ACS national household income proportions
# Broad categories based on Census ACS B19001
national_income_df <- data.frame(
  Income = c(
    "Less than $25,000",
    "$25,000 to less than $50,000",
    "$50,000 to less than $75,000",
    "$75,000 or more"
  ),
  Expected_Prop = c(
    0.161,
    0.171,
    0.148,
    0.520
  )
) %>%
  mutate(Expected_Prop = Expected_Prop / sum(Expected_Prop))

# Clean BRFSS veteran income data
brfss_veterans_income <- brfss %>%
  filter(veteran == "Yes", !is.na(income)) %>%
  mutate(
    income_chr = as.character(income),
    income_group = case_when(
      income_chr %in% c(
        "Less than $10000",
        "$10000 to $14999",
        "$15000 to $19999",
        "$20000 to $24999"
      ) ~ "Less than $25,000",
      
      income_chr %in% c(
        "$25000 to $34999",
        "$35000 to $49999"
      ) ~ "$25,000 to less than $50,000",
      
      income_chr == "$50000 to $74999" ~ "$50,000 to less than $75,000",
      
      income_chr %in% c(
        "$75000 to $99999",
        "$100000 to $149999",
        "$150000 to $199999",
        "$200000 or more"
      ) ~ "$75,000 or more",
      
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(income_group))

# Observed veteran income counts
observed_income_counts <- table(brfss_veterans_income$income_group)

# Expected proportions ordered to match observed table
expected_income_props_ordered <- national_income_df$Expected_Prop[
  match(names(observed_income_counts), national_income_df$Income)
]

expected_income_props_ordered <- expected_income_props_ordered / sum(expected_income_props_ordered)

# Chi-square goodness-of-fit test
income_chisq_test <- chisq.test(
  observed_income_counts,
  p = expected_income_props_ordered
)

income_chisq_test

# Table showing which income groups differ most
income_diff_table <- data.frame(
  Income = names(observed_income_counts),
  Observed_Count = as.numeric(observed_income_counts),
  Expected_Prop = expected_income_props_ordered
) %>%
  mutate(
    Total = sum(Observed_Count),
    Observed_Prop = Observed_Count / Total,
    Expected_Count = Expected_Prop * Total,
    Difference = Observed_Prop - Expected_Prop,
    ChiSq_Contribution = (Observed_Count - Expected_Count)^2 / Expected_Count,
    Direction = ifelse(Difference > 0, "Overrepresented", "Underrepresented")
  ) %>%
  arrange(desc(ChiSq_Contribution))

income_diff_table

# One-sample proportion tests with Bonferroni correction
income_prop_tests <- data.frame(
  Income = names(observed_income_counts),
  Observed_Count = as.numeric(observed_income_counts),
  Expected_Prop = expected_income_props_ordered
) %>%
  mutate(
    Total = sum(Observed_Count),
    Observed_Prop = Observed_Count / Total,
    Difference = Observed_Prop - Expected_Prop
  ) %>%
  rowwise() %>%
  mutate(
    p_value = prop.test(
      x = Observed_Count,
      n = Total,
      p = Expected_Prop,
      correct = FALSE
    )$p.value
  ) %>%
  ungroup() %>%
  mutate(
    p_bonferroni = p.adjust(p_value, method = "bonferroni"),
    Direction = ifelse(Difference > 0, "Overrepresented", "Underrepresented")
  ) %>%
  select(
    Income,
    Observed_Prop,
    Expected_Prop,
    p_bonferroni,
    Direction
  ) %>%
  arrange(p_bonferroni)

income_prop_tests
# Plot observed vs expected proportions using same color scheme
income_diff_table %>%
  select(Income, Observed_Prop, Expected_Prop) %>%
  pivot_longer(
    cols = c(Observed_Prop, Expected_Prop),
    names_to = "Type",
    values_to = "Proportion"
  ) %>%
  mutate(
    Type = recode(
      Type,
      "Observed_Prop" = "Observed",
      "Expected_Prop" = "Expected"
    )
  ) %>%
  ggplot(aes(x = Income, y = Proportion, fill = Type)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(
    values = c("Expected" = vet_blue, "Observed" = vet_teal)
  ) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Veteran Income Distribution vs 2022 U.S. National Average",
    x = "Income Level",
    y = "Proportion",
    fill = "Group"
  ) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```




# Veteran vs Non Veteran: Mental + Physical Health



```{r}
#| echo: false
library(dplyr)
brfss = read.csv("brfss2022_m62.csv", stringsAsFactors=TRUE)
veteran <- subset(brfss, veteran == "Yes")
nrow(veteran)
nonVeteran <- subset(brfss, veteran == "No")
nrow(nonVeteran)
```

Comparisons of physical health between veterans and non-veterans:

Arthritis: p-value $< 2.2 \times 10^{-16}$
```{r}
#| echo: false

veteranArthClean <- veteran %>% filter(!is.na(arthritis))
nonVeteranArthClean <- nonVeteran %>% filter(!is.na(arthritis))

arthritisVeterans <- sum(veteranArthClean$arthritis == "Yes")
arthritisNonVeterans <- sum(nonVeteranArthClean$arthritis == "Yes")

successes <- c(arthritisVeterans, arthritisNonVeterans)
n_obs <- c(nrow(veteranArthClean), nrow(nonVeteranArthClean))

arthritis_prop_test <- prop.test(successes, n_obs)
arthritis_prop_test
```

Heart Disease: p-value $< 2.2 \times 10^{-16}$
```{r}
#| echo: false

veteranHDClean <- veteran %>% filter(!is.na(heartdisease))
nonVeteranHDClean <- nonVeteran %>% filter(!is.na(heartdisease))

hdVeterans <- sum(veteranHDClean$heartdisease == "Yes")
hdNonVeterans <- sum(nonVeteranHDClean$heartdisease == "Yes")

successes <- c(hdVeterans, hdNonVeterans)
n_obs <- c(nrow(veteranHDClean), nrow(nonVeteranHDClean))

hd_prop_test <- prop.test(successes, n_obs)
hd_prop_test
```

Cancer: p-value $< 2.2 \times 10^{-16}$
```{r}
#| echo: false

veteranCancerClean <- veteran %>% filter(!is.na(cancer))
nonVeteranCancerClean <- nonVeteran %>% filter(!is.na(cancer))

cancerVeterans <- sum(veteranCancerClean$cancer == "Yes")
cancerNonVeterans <- sum(nonVeteranCancerClean$cancer == "Yes")

successes <- c(cancerVeterans, cancerNonVeterans)
n_obs <- c(nrow(veteranCancerClean), nrow(nonVeteranCancerClean))

cancer_prop_test <- prop.test(successes, n_obs)
cancer_prop_test
```

Diabetes: p-value $< 2.2 \times 10^{-16}$
```{r}
#| echo: false

veteranDiabetesClean <- veteran %>% filter(!is.na(diabetes))
nonVeteranDiabetesClean <- nonVeteran %>% filter(!is.na(diabetes))

diabetesVeterans <- sum(veteranDiabetesClean$diabetes == "Yes")
diabetesNonVeterans <- sum(nonVeteranDiabetesClean$diabetes == "Yes")

successes <- c(diabetesVeterans, diabetesNonVeterans)
n_obs <- c(nrow(veteranDiabetesClean), nrow(nonVeteranDiabetesClean))

diabetes_prop_test <- prop.test(successes, n_obs)
diabetes_prop_test
```

COPD: p-value $< 2.2 \times 10^{-16}$
```{r}
#| echo: false

veteranCOPDClean <- veteran %>% filter(!is.na(copd))
nonVeteranCOPDClean <- nonVeteran %>% filter(!is.na(copd))

copdVeterans <- sum(veteranCOPDClean$copd == "Yes")
copdNonVeterans <- sum(nonVeteranCOPDClean$copd == "Yes")

successes <- c(copdVeterans, copdNonVeterans)
n_obs <- c(nrow(veteranCOPDClean), nrow(nonVeteranCOPDClean))

copd_prop_test <- prop.test(successes, n_obs)
copd_prop_test
```

daysphyshlthnotgood: p-value $< 2.2 \times 10^{-16}$
```{r}
#| echo: false

veteranDaysphyshlthnotgoodClean <- veteran %>% filter(!is.na(daysphyshlthnotgood))
nonVeteranDaysphyshlthnotgoodClean <- nonVeteran %>% filter(!is.na(daysphyshlthnotgood))

daysphyshlthnotgoodVeterans <- mean(veteranDaysphyshlthnotgoodClean$daysphyshlthnotgood)
daysphyshlthnotgoodNonVeterans <- mean(nonVeteranDaysphyshlthnotgoodClean$daysphyshlthnotgood)

daysphyshlthnotgood_t_test <- t.test(veteranDaysphyshlthnotgoodClean$daysphyshlthnotgood,
                                     nonVeteranDaysphyshlthnotgoodClean$daysphyshlthnotgood,
                                     alternative="two.sided",
                                     mu=0, paired=FALSE, var.equal = FALSE,
                                     conf.level=0.95)
daysphyshlthnotgood_t_test
```

Check Assumptions (equal variances):
```{r}
#| echo: false
#| 
hist(veteranDaysphyshlthnotgoodClean$daysphyshlthnotgood,
     main = "Days Physical Health Not Good (Veteran)",
     xlab = "Number of days in past month physical health was not good")
hist(nonVeteranDaysphyshlthnotgoodClean$daysphyshlthnotgood,
     main = "Days Physical Health Not Good (Non-Veteran)",
     xlab = "Number of days in past month physical health was not good")
```


```{r}
#| echo: false
# Summary of physical health tests

factors <- c("Arthritis", "Heart Disease", "Cancer", "Diabetes", "COPD", "Days Physical Health Not Good")
prop_vet <- c(0.4356422, 0.1244373, 0.1789827, 0.1962027, 0.11969946, 4.972634)
prop_nonVet <- c(0.3455287, 0.0543102, 0.1105903, 0.1340290, 0.07724301, 4.338173)
p_values <- rep("< 2.2e-16", 6)

physical_tests <- data.frame(
  Factors = factors,
  Veteran = prop_vet,
  NonVeteran = prop_nonVet,
  pvalue = p_values
)

physical_tests
```

```{r}
#| echo: false
# Enhanced R Plotting Script
#| echo: false

# --- Data ---
labels_a <- c("Arthritis", "Heart Disease", "Cancer", "Diabetes", "COPD")
prop_vet <- c(0.4356422, 0.1244373, 0.1789827, 0.1962027, 0.11969946)
prop_nonVet <- c(0.3455287, 0.0543102, 0.1105903, 0.1340290, 0.07724301)
data_matrix_a <- rbind(prop_vet, prop_nonVet)
colnames(data_matrix_a) <- labels_a

labels_b <- c("Days Phys. Health\nNot Good")
mean_vet <- c(4.972634)
mean_nonVet <- c(4.338173)
data_matrix_b <- rbind(mean_vet, mean_nonVet)
colnames(data_matrix_b) <- labels_b

chart_colors <- c("#2a79d5", "#29b9d6")

# --- Layout Configuration ---
par(mfrow = c(1, 2), mar = c(5, 4, 4, 2))

# --- Plot 1: Chronic Conditions ---
# Initialize the plot area without drawing bars yet to set up the grid
bp1 <- barplot(data_matrix_a, beside = TRUE, col = NA, border = NA, 
               main = "Chronic Health Conditions", ylab = "Proportion", 
               ylim = c(0, 0.5), las = 1, cex.names = 0.7, font.main = 2)

# Draw the grid lines FIRST
grid(nx = NA, ny = NULL, col = "gray90", lty = "solid")

# Overlay the bars ON TOP of the grid
barplot(data_matrix_a, 
        beside = TRUE, 
        col = chart_colors,
        border = "white",
        add = TRUE,          # This tells R to draw over the existing grid
        las = 1, 
        cex.names = 0.7,
        xaxt = "n")          # Prevents double-drawing labels

legend("topright", legend = c("Veteran", "Non-Veteran"), 
       fill = chart_colors, bty = "n", cex = 0.7)

# --- Plot 2: Quality of Life ---
barplot(data_matrix_b, beside = TRUE, col = NA, border = NA, 
        main = "Average Days Unwell", ylab = "Mean Days", 
        ylim = c(0, 6), las = 1, cex.names = 0.7, font.main = 2)

grid(nx = NA, ny = NULL, col = "gray90", lty = "solid")

barplot(data_matrix_b, 
        beside = TRUE, 
        col = chart_colors,
        border = "white",
        add = TRUE, 
        las = 2, 
        cex.names = 0.7,
        xaxt = "n")

# Reset
par(mfrow = c(1, 1))
```

Comparisons of mental health between veterans and non-veterans:

Depression: p-value $< 2.2 \times 10^{-16}$
```{r}
#| echo: false

veteranDepressClean <- veteran %>% filter(!is.na(depress))
nonVeteranDepressClean <- nonVeteran %>% filter(!is.na(depress))

depressVeterans <- sum(veteranDepressClean$depress == "Yes")
depressNonVeterans <- sum(nonVeteranDepressClean$depress == "Yes")

successes <- c(depressVeterans, depressNonVeterans)
n_obs <- c(nrow(veteranDepressClean), nrow(nonVeteranDepressClean))

depress_prop_test <- prop.test(successes, n_obs)
depress_prop_test
```

daysmenthlthnotgood: p-value $< 2.2 \times 10^{-16}$
```{r}
#| echo: false

veteranDaysmenthlthnotgoodClean <- veteran %>% filter(!is.na(daysmenthlthnotgood))
nonVeteranDaysmenthlthnotgoodClean <- nonVeteran %>% filter(!is.na(daysmenthlthnotgood))

daysmenthlthnotgoodVeterans <- mean(veteranDaysmenthlthnotgoodClean$daysmenthlthnotgood)
daysmenthlthnotgoodNonVeterans <- mean(nonVeteranDaysmenthlthnotgoodClean$daysmenthlthnotgood)

daysmenthlthnotgood_t_test <- t.test(veteranDaysmenthlthnotgoodClean$daysmenthlthnotgood,
                                     nonVeteranDaysmenthlthnotgoodClean$daysmenthlthnotgood,
                                     alternative="two.sided",
                                     mu=0, paired=FALSE, var.equal = TRUE,
                                     conf.level=0.95)
daysmenthlthnotgood_t_test
```

Check Assumptions (equal variances):
```{r}
#| echo: false
#| 
hist(veteranDaysmenthlthnotgoodClean$daysmenthlthnotgood,
     main = "Days Mental Health Not Good (Veteran)",
     xlab = "Number of days in past month mental health was not good")
hist(nonVeteranDaysmenthlthnotgoodClean$daysmenthlthnotgood,
     main = "Days Mental Health Not Good (Non-Veteran)",
     xlab = "Number of days in past month mental health was not good")
```

```{r}
#| echo: false
# Summary of Mental Health Tests

factors <- c("Depression", "Days Physical Health Not Good")
prop_vet <- c(0.1757344, 3.517690)
prop_nonVet <- c(0.2166535, 4.462462)
p_values <- rep("< 2.2e-16", 2)

mental_tests <- data.frame(
  Factors = factors,
  Veteran = prop_vet,
  NonVeteran = prop_nonVet,
  pvalue = p_values
)

mental_tests
```

```{r}
#| echo: false
# generated with help from gemini

# Your vectors
labels_a <- c("Depression")
prop_vet <- c(0.1757344)
prop_nonVet <- c(0.2166535)

# Combine into a matrix
data_matrix_a <- rbind(prop_vet, prop_nonVet)

# Assign column names to the matrix
colnames(data_matrix_a) <- labels_a

# Your vectors
labels_b <- c("daysmenthlthnotgood")
mean_vet <- c(3.517690)
mean_nonVet <- c(4.462462)

# Combine into a matrix
data_matrix_b <- rbind(mean_vet, mean_nonVet)

# Assign column names to the matrix
colnames(data_matrix_b) <- labels_b

# Create the grouped barplot
# Adjust margins so the long label fits (bottom, left, top, right)
par(mar = c(12, 4, 4, 2) + 0.1)
par(mfrow=c(1,2))

barplot(data_matrix_a, 
        beside = TRUE,                   # This places the bars side-by-side
        col = c("skyblue", "salmon"),    # Colors for the two groups
        legend.text = c("Veteran", "Non-Veteran"),
        args.legend = list(x = "topright", bty = "n"),
        main = "Health Indicators: ",
        ylab = "Value",
        las = 1,                         # Rotates x-axis labels to be vertical
        cex.names = 0.8)                 # Shrinks the label text slightly

barplot(data_matrix_b, 
        beside = TRUE,                   # This places the bars side-by-side
        col = c("skyblue", "salmon"),    # Colors for the two groups
        legend.text = c("Veteran", "Non-Veteran"),
        args.legend = list(x = "topright", bty = "n"),
        main = "Veterans vs Non-Veterans",
        ylab = "Value",
        las = 1,                         # Rotates x-axis labels to be vertical
        cex.names = 0.8)                 # Shrinks the label text slightly
```

What is the difference in the probability of having a specific health factor being a veteran vs. non-veteran?

```{r}
#| echo: false

arthritisClean <- brfss %>% filter(!is.na(arthritis))

arthritisClean$veteran <- as.factor(arthritisClean$veteran)

arthritisClean$arthritis <- as.factor(arthritisClean$arthritis)

arthritisMod <- glm(formula = arthritis ~ veteran, data = arthritisClean, family = "binomial")

summary(arthritisMod)
```

```{r}
p_vet_arthritis <- plogis(-0.63875 + 0.37988 * 1)
p_nonVet_arthritis <- plogis(-0.63875 + 0.37988 * 0)

cat("P(Arthritis | Veteran) = ", p_vet_arthritis)
cat('\n')
cat("P(Arthritis | Non-Veteran) = ", p_nonVet_arthritis)
```

```{r}
#| echo: false

hdClean <- brfss %>% filter(!is.na(heartdisease))

hdClean$veteran <- as.factor(hdClean$veteran)

hdClean$heartdisease <- as.factor(hdClean$heartdisease)

hdMod <- glm(formula = heartdisease ~ veteran, data = hdClean, family = "binomial")

summary(hdMod)
```

```{r}
p_vet_hd <- plogis(-2.857202 + 0.906138 * 1)
p_nonVet_hd <- plogis(-2.857202 + 0.906138 * 0)

cat("P(Heart Disease | Veteran) = ", p_vet_hd)
cat('\n')
cat("P(Heart Disease | Non-Veteran) = ", p_nonVet_hd)
```

```{r}
#| echo: false

cancerClean <- brfss %>% filter(!is.na(cancer))

cancerClean$veteran <- as.factor(cancerClean$veteran)

cancerClean$cancer <- as.factor(cancerClean$cancer)

cancerMod <- glm(formula = cancer ~ veteran, data = cancerClean, family = "binomial")

summary(cancerMod)
```

```{r}
p_vet_cancer <- plogis(-2.084725 + 0.56147 * 1)
p_nonVet_cancer <- plogis(-2.084725 + 0.56147 * 0)

cat("P(Cancer | Veteran) = ", p_vet_cancer)
cat('\n')
cat("P(Cancer | Non-Veteran) = ", p_nonVet_cancer)
```

```{r}
#| echo: false

diabetesClean <- brfss %>% filter(!is.na(diabetes))

diabetesClean$veteran <- as.factor(diabetesClean$veteran)

diabetesClean$diabetes <- as.factor(diabetesClean$diabetes)

diabetesMod <- glm(formula = diabetes ~ veteran, data = diabetesClean, family = "binomial")

summary(diabetesMod)
```

```{r}
p_vet_diabetes <- plogis(-1.602633 + 0.370480 * 1)
p_nonVet_diabetes <- plogis(-1.602633 + 0.370480 * 0)

cat("P(Diabetes | Veteran) = ", p_vet_diabetes)
cat('\n')
cat("P(Diabetes | Non-Veteran) = ", p_nonVet_diabetes)
```

```{r}
#| echo: false

copdClean <- brfss %>% filter(!is.na(copd))

copdClean$veteran <- as.factor(copdClean$veteran)

copdClean$diabetes <- as.factor(copdClean$copd)

copdMod <- glm(formula = copd ~ veteran, data = copdClean, family = "binomial")

summary(copdMod)
```

```{r}
p_vet_copd <- plogis(-2.480409 + 0.485130 * 1)
p_nonVet_copd <- plogis(-2.480409 + 0.485130 * 0)

cat("P(COPD | Veteran) = ", p_vet_copd)
cat('\n')
cat("P(COPD | Non-Veteran) = ", p_nonVet_copd)
```

```{r}
#| echo: false

depressClean <- brfss %>% filter(!is.na(depress))

depressClean$veteran <- as.factor(depressClean$veteran)

depressClean$depress <- as.factor(depressClean$depress)

depressMod <- glm(formula = depress ~ veteran, data = depressClean, family = "binomial")

summary(depressMod)
```

```{r}
p_vet_depress <- plogis(-1.285276 - 0.260243 * 1)
p_nonVet_depress <- plogis(-1.285276 - 0.260243 * 0)

cat("P(Depressive Disorder Diagnosis | Veteran) = ", p_vet_depress)
cat('\n')
cat("P(Depressive Disorder Diagnosis | Non-Veteran) = ", p_nonVet_depress)
```



What are some factors that influence physical health among veterans

```{r}
library(sjPlot)
library(sjlabelled)
library(sjmisc)
library(ggplot2)
brfss = read.csv("brfss2022_m62.csv", stringsAsFactors=TRUE)

head(brfss)
```

```{r Physical Health: Heart Disease}
#| echo: false
veteran <- subset(brfss, veteran == "Yes")



# Code below is Gemini generated: Prompt was that given these age groups, merge them so that we have three columns of young adult, middle age, and seniors

# 1. Create the new categorization logic
# We assign the categories based on the modified strings
veteran$age_category <- ifelse(veteran$age %in% c("25 to 29", "30 to 34", "35 to 39"), "young adult",
                        ifelse(veteran$age %in% c("40 to 44", "45 to 49", "50 to 54", "55 to 59", "60 to 64"), "middle-aged",
                        ifelse(veteran$age %in% c("65 to 69", "70 to 74", "75 to 79", "80 or older"), "senior", NA)))

# 2. Remove the old granular column and rename the new one to 'age'
veteran$age <- NULL
names(veteran)[names(veteran) == "age_category"] <- "age"

#Code below is Gemini generated: Prompt was that given these income groups, merge them so that we have three columns of high income, medium income, low income
# 1. Create the new categorization logic
veteran$income <- ifelse(veteran$income %in% c("$150000 to $199999", "$200000 or more"), 
                                "high income",
                         ifelse(veteran$income %in% c("Less than $10000", "$10000 to $14999", 
                                                             "$15000 to $19999", "$20000 to $24999", 
                                                             "$25000 to $34999"), 
                                "low income",
                         ifelse(is.na(veteran$income), NA, "medium income")))



```


```{r}
# Lines below generated by Gemini, prompt was that given these columns, drop any rows where any of these columns are NA, and only keep entries in diabetes that's "Yes" or "No"
# Define the variables of interest
vars <- c("heartdisease", "sex", "bmi", "cancer", 
          "daysphyshlthnotgood", "anyexer", "race", 
          "age", "bingedrinker", "weeklydrinks", "heartdisease", "income")

# Subset rows where all specified columns are non-NA
veteran_physical <- veteran[complete.cases(veteran[, vars]), ]

# This keeps only rows where diabetes is exactly "Yes" or "No"
veteran_physical <- veteran_physical[veteran_physical$diabetes %in% c("Yes", "No"), ]

#The lines below are generated by Gemini, prompt was that given these column names, show how many veterans is in each category
# 1. Define the variables
vars <- c("daysmenthlthnotgood", "income", "race", "age", "bingedrinker", 
          "weeklydrinks", "sex", "anyexer", "depress", "bmi", 
          "cancer", "diabetes", "daysphyshlthnotgood")

# 2. Calculate the count of non-NA values for each column
# !is.na(x) returns TRUE for non-missing values; colSums totals those TRUEs
non_na_counts <- colSums(!is.na(veteran_physical[, vars]))

# 3. View the results as a data frame for readability
as.data.frame(non_na_counts)

```


```{r}
#| echo: false
model_heartdisease <- glm(formula = heartdisease ~ sex + bmi + income +
                 daysphyshlthnotgood + anyexer + race +
                 age + bingedrinker + weeklydrinks , data=veteran_physical, family=binomial)
```



```{r Physical Health: Cancer}
#| echo: false

model_cancer <- glm(formula = cancer ~ sex + bmi +
                 daysphyshlthnotgood + anyexer + race +
                 age + bingedrinker + weeklydrinks + income, data=veteran_physical, family=binomial)


```


```{r Physical Health: Diabetes}
#| echo: false

model_diabetes <- glm(formula = diabetes ~ sex + bmi + income +
                 daysphyshlthnotgood + anyexer + race +
                 age + bingedrinker + weeklydrinks , data=veteran_physical, family=binomial)

```

```{r Physical Health: BMI}
#| echo: false

model_bmi <- lm(formula = bmi ~ sex + income +
                 daysphyshlthnotgood + anyexer + race +
                 age + bingedrinker + weeklydrinks , data=veteran_physical)


```

What are some factors that influence mental health among veterans

```{r Mental Health: Depression}
#| echo: false

# Two lines below generated by Gemini, prompt was that given these columns, drop any rows where any of these columns are NA
# Define the variables of interest

vars_mental <- c("daysmenthlthnotgood", "income", "race", "age", "bingedrinker", "weeklydrinks", "sex", "anyexer", "depress")

# Subset rows where all specified columns are non-NA
veteran_mental <- veteran[complete.cases(veteran[, vars_mental]), ]


model_depression <- glm(formula = depress ~ sex + income +
                 daysmenthlthnotgood + anyexer + race +
                 age + bingedrinker + weeklydrinks , data=veteran_mental, family=binomial)

```


```{r}
#Below code generated by Gemini, instructions was to make a data frame table that included the below columns based on a given regression model

# 1. Updated Helper Function to include p-values
get_sig_df <- function(model) {
  # Extract coefficients into a data frame
  stats <- as.data.frame(summary(model)$coefficients)
  
  # Standardize column names
  colnames(stats) <- c("Coefficient", "StdError", "t_z_value", "p_value")
  
  # Move Predictors to a column
  stats$Predictor <- rownames(stats)
  
  # Calculate Odds Ratio
  stats$Odds_Ratio <- exp(stats$Coefficient)
  
  # Filter for significance (p < 0.01) to be on the safe side
  # Select Predictor, Coefficient, Odds_Ratio, and p_value
  #final_df <- stats[stats$p_value < 0.01, c("Predictor", "Coefficient","p_value", "Odds_Ratio")]
  
  adjusted_p_value <- 0.05/9
  
  final_df <- stats[stats$p_value < adjusted_p_value, c("Predictor", "Coefficient","p_value", "Odds_Ratio")]
  
  # Reset row names
  rownames(final_df) <- NULL
  
  return(final_df)
}

# 2. Generate the 5 data frames
sig_heart    <- get_sig_df(model_heartdisease)
sig_cancer   <- get_sig_df(model_cancer)
sig_diabetes <- get_sig_df(model_diabetes)
sig_bmi      <- get_sig_df(model_bmi)
sig_depress  <- get_sig_df(model_depression)


# 3. Display the tables
cat("\n--- Table 1: Heart Disease ---\n")
print(sig_heart)

cat("\n--- Table 2: Cancer ---\n")
print(sig_cancer)

cat("\n--- Table 3: Diabetes ---\n")
print(sig_diabetes)

cat("\n--- Table 4: BMI ---\n")
print(sig_bmi)

cat("\n--- Table 5: Depression ---\n")
print(sig_depress)

```


```{r}
plot(model_bmi, which = 1)
plot(model_bmi, which = 2)


```





