---
title: "CDC BRFSS Analysis: Veteran Health"

author: "Math 62 - Spring 2026"
format:
  pdf:
    knitr:
      opts_chunk: 
        out-width: "60%"
        fig-align: "center"
        size: "scriptsize"
    fontsize: 10pt
    toc: false
    toc-params: "allowframebreaks"
    include-in-header:
      - text: |
          \usepackage{etoolbox}
          \usepackage{hyperref}
          
execute:
  echo: false
  warning: false
  message: false
  eval: true

fig-align: center
fig-height: 6
fig-width: 8

---

# Team Members

Li, Elsa

Louie, Amanda

Sharma, Anika

---

```{r}
#| echo: true
2 * 2
```

The `echo: true` option enables the printing of code (only output is displayed).

Amanda Section 

```{r}
#| echo: false
library(dplyr)
brfss = read.csv("brfss2022_m62.csv", stringsAsFactors=TRUE)
veteran <- subset(brfss, veteran == "Yes")
nrow(veteran)
nonVeteran <- subset(brfss, veteran == "No")
nrow(nonVeteran)
```

Comparisons of physical health between veterans and non-veterans:

Arthritis: p-value $< 2.2 \times 10^{-16}$
```{r}
#| echo: false

veteranArthClean <- veteran %>% filter(!is.na(arthritis))
nonVeteranArthClean <- nonVeteran %>% filter(!is.na(arthritis))

arthritisVeterans <- sum(veteranArthClean$arthritis == "Yes")
arthritisNonVeterans <- sum(nonVeteranArthClean$arthritis == "Yes")

successes <- c(arthritisVeterans, arthritisNonVeterans)
n_obs <- c(nrow(veteranArthClean), nrow(nonVeteranArthClean))

arthritis_prop_test <- prop.test(successes, n_obs)
arthritis_prop_test
```

Heart Disease: p-value $< 2.2 \times 10^{-16}$
```{r}
#| echo: false

veteranHDClean <- veteran %>% filter(!is.na(heartdisease))
nonVeteranHDClean <- nonVeteran %>% filter(!is.na(heartdisease))

hdVeterans <- sum(veteranHDClean$heartdisease == "Yes")
hdNonVeterans <- sum(nonVeteranHDClean$heartdisease == "Yes")

successes <- c(hdVeterans, hdNonVeterans)
n_obs <- c(nrow(veteranHDClean), nrow(nonVeteranHDClean))

hd_prop_test <- prop.test(successes, n_obs)
hd_prop_test
```

Cancer: p-value $< 2.2 \times 10^{-16}$
```{r}
#| echo: false

veteranCancerClean <- veteran %>% filter(!is.na(cancer))
nonVeteranCancerClean <- nonVeteran %>% filter(!is.na(cancer))

cancerVeterans <- sum(veteranCancerClean$cancer == "Yes")
cancerNonVeterans <- sum(nonVeteranCancerClean$cancer == "Yes")

successes <- c(cancerVeterans, cancerNonVeterans)
n_obs <- c(nrow(veteranCancerClean), nrow(nonVeteranCancerClean))

cancer_prop_test <- prop.test(successes, n_obs)
cancer_prop_test
```

Diabetes: p-value $< 2.2 \times 10^{-16}$
```{r}
#| echo: false

veteranDiabetesClean <- veteran %>% filter(!is.na(diabetes))
nonVeteranDiabetesClean <- nonVeteran %>% filter(!is.na(diabetes))

diabetesVeterans <- sum(veteranDiabetesClean$diabetes == "Yes")
diabetesNonVeterans <- sum(nonVeteranDiabetesClean$diabetes == "Yes")

successes <- c(diabetesVeterans, diabetesNonVeterans)
n_obs <- c(nrow(veteranDiabetesClean), nrow(nonVeteranDiabetesClean))

diabetes_prop_test <- prop.test(successes, n_obs)
diabetes_prop_test
```

COPD: p-value $< 2.2 \times 10^{-16}$
```{r}
#| echo: false

veteranCOPDClean <- veteran %>% filter(!is.na(copd))
nonVeteranCOPDClean <- nonVeteran %>% filter(!is.na(copd))

copdVeterans <- sum(veteranCOPDClean$copd == "Yes")
copdNonVeterans <- sum(nonVeteranCOPDClean$copd == "Yes")

successes <- c(copdVeterans, copdNonVeterans)
n_obs <- c(nrow(veteranCOPDClean), nrow(nonVeteranCOPDClean))

copd_prop_test <- prop.test(successes, n_obs)
copd_prop_test
```

daysphyshlthnotgood: p-value $< 2.2 \times 10^{-16}$
```{r}
#| echo: false

veteranDaysphyshlthnotgoodClean <- veteran %>% filter(!is.na(daysphyshlthnotgood))
nonVeteranDaysphyshlthnotgoodClean <- nonVeteran %>% filter(!is.na(daysphyshlthnotgood))

daysphyshlthnotgoodVeterans <- mean(veteranDaysphyshlthnotgoodClean$daysphyshlthnotgood)
daysphyshlthnotgoodNonVeterans <- mean(nonVeteranDaysphyshlthnotgoodClean$daysphyshlthnotgood)

daysphyshlthnotgood_t_test <- t.test(veteranDaysphyshlthnotgoodClean$daysphyshlthnotgood,
                                     nonVeteranDaysphyshlthnotgoodClean$daysphyshlthnotgood,
                                     alternative="two.sided",
                                     mu=0, paired=FALSE, var.equal = FALSE,
                                     conf.level=0.95)
daysphyshlthnotgood_t_test
```

Comparisons of mental health between veterans and non-veterans:

Depression: p-value $< 2.2 \times 10^{-16}$
```{r}
#| echo: false

veteranDepressClean <- veteran %>% filter(!is.na(depress))
nonVeteranDepressClean <- nonVeteran %>% filter(!is.na(depress))

depressVeterans <- sum(veteranDepressClean$depress == "Yes")
depressNonVeterans <- sum(nonVeteranDepressClean$depress == "Yes")

successes <- c(depressVeterans, depressNonVeterans)
n_obs <- c(nrow(veteranDepressClean), nrow(nonVeteranDepressClean))

depress_prop_test <- prop.test(successes, n_obs)
depress_prop_test
```

daysmenthlthnotgood: p-value $< 2.2 \times 10^{-16}$
```{r}
#| echo: false

veteranDaysmenthlthnotgoodClean <- veteran %>% filter(!is.na(daysmenthlthnotgood))
nonVeteranDaysmenthlthnotgoodClean <- nonVeteran %>% filter(!is.na(daysmenthlthnotgood))

daysmenthlthnotgoodVeterans <- mean(veteranDaysmenthlthnotgoodClean$daysmenthlthnotgood)
daysmenthlthnotgoodNonVeterans <- mean(nonVeteranDaysmenthlthnotgoodClean$daysmenthlthnotgood)

daysmenthlthnotgood_t_test <- t.test(veteranDaysmenthlthnotgoodClean$daysmenthlthnotgood,
                                     nonVeteranDaysmenthlthnotgoodClean$daysmenthlthnotgood,
                                     alternative="two.sided",
                                     mu=0, paired=FALSE, var.equal = FALSE,
                                     conf.level=0.95)
daysmenthlthnotgood_t_test
```

Elsa Section

What are some factors that influence physical health among veterans

```{r}
library(sjPlot)
library(sjlabelled)
library(sjmisc)
library(ggplot2)
brfss = read.csv("brfss2022_m62.csv", stringsAsFactors=TRUE)

head(brfss)
```

```{r Physical Health: Heart Disease}
#| echo: false
veteran <- subset(brfss, veteran == "Yes")


# Code below is Gemini generated: Prompt was that given these age groups, merge them so that we have three columns of young adult, middle age, and seniors

# 1. Create the new categorization logic
# We assign the categories based on the modified strings
veteran$age_category <- ifelse(veteran$age %in% c("25 to 29", "30 to 34", "35 to 39"), "young adult",
                        ifelse(veteran$age %in% c("40 to 44", "45 to 49", "50 to 54", "55 to 59", "60 to 64"), "middle-aged",
                        ifelse(veteran$age %in% c("65 to 69", "70 to 74", "75 to 79", "80 or older"), "senior", NA)))

# 2. Remove the old granular column and rename the new one to 'age'
veteran$age <- NULL
names(veteran)[names(veteran) == "age_category"] <- "age"

```

```{r}
# Two lines below generated by Gemini, prompt was that given these columns, drop any rows where any of these columns are NA
# Define the variables of interest
vars <- c("heartdisease", "sex", "bmi", "cancer", "copd", 
          "daysphyshlthnotgood", "anyexer", "race", 
          "age", "bingedrinker", "weeklydrinks")

# Subset rows where all specified columns are non-NA
veteran_physical <- veteran[complete.cases(veteran[, vars]), ]


model_heartdiease <- glm(formula = heartdisease ~ sex + bmi + cancer + copd +
                 daysphyshlthnotgood + anyexer + race +
                 age + bingedrinker + weeklydrinks , data=veteran_physical, family=binomial)

#The line below are Gemini generated: Prompt was that given a glm model, create automatic stepwise param selection
# Perform stepwise selection using AIC
# direction = "both" checks for variables to add AND remove at each step
final_model_heartdisease <- step(model_heartdiease, direction = "both")




```
```{r}
# This line is Gemini Generated: Prompt was that given a glm model, create automatic stepwise param selection
# Perform stepwise selection using AIC
# View the final selected model
summary(final_model_heartdisease)
```

```{r}
theme_set(theme_sjplot())
tab_model(final_model_heartdisease, show.stat = FALSE)
```


```{r Physical Health: Cancer}
#| echo: false

model_cancer <- glm(formula = cancer ~ sex + bmi  + copd +
                 daysphyshlthnotgood + anyexer + race +
                 age + bingedrinker + weeklydrinks , data=veteran_physical, family=binomial)

#The two lines below are Gemini generated: Prompt was that given a glm model, create automatic stepwise param selection
# Perform stepwise selection using AIC
# direction = "both" checks for variables to add AND remove at each step
final_model_cancer <- step(model_cancer)

# View the final selected model
summary(final_model_cancer)
tab_model(final_model_cancer, show.stat = FALSE)
```


```{r Physical Health: Diabetes}
#| echo: false

model_diabetes <- glm(formula = diabetes ~ sex + bmi + cancer + copd +
                 daysphyshlthnotgood + anyexer + race +
                 age + bingedrinker + weeklydrinks , data=veteran_physical, family=binomial)

#The two lines below are Gemini generated: Prompt was that given a glm model, create automatic stepwise param selection
# Perform stepwise selection using AIC
# direction = "both" checks for variables to add AND remove at each step
final_model_diabetes <- step(model_diabetes)

# View the final selected model
summary(final_model_diabetes)
tab_model(final_model_diabetes, show.stat = FALSE)
```
What are some factors that influence mental health among veterans

```{r Mental Health: Depression}
#| echo: false

# Two lines below generated by Gemini, prompt was that given these columns, drop any rows where any of these columns are NA
# Define the variables of interest

vars_mental <- c("daysmenthlthnotgood", "medcost", "income", "race", "age", "bingedrinker", "weeklydrinks", "sex", "depress")

# Subset rows where all specified columns are non-NA
veteran_mental <- veteran[complete.cases(veteran[, vars_mental]), ]


model_depression <- glm(formula = depress ~ sex + medcost + income + race +
                 age + bingedrinker + weeklydrinks + daysmenthlthnotgood , data=veteran_mental, family=binomial)

#The two lines below are Gemini generated: Prompt was that given a glm model, create automatic stepwise param selection
# Perform stepwise selection using AIC
# direction = "both" checks for variables to add AND remove at each step
final_model_depression <- step(model_depression, direction = "both")

# View the final selected model
summary(final_model_depression)
tab_model(final_model_depression, show.stat = FALSE)
```


```{r Mental Health: Daysmenthlthnotgood}
#| echo: false

# Two lines below generated by Gemini, prompt was that given these columns, drop any rows where any of these columns are NA
# Define the variables of interest
model_daysmenthlthnotgood<- lm(formula = daysmenthlthnotgood ~ sex + medcost + income + race +
                 age + bingedrinker + weeklydrinks + depress , data=veteran_mental)

#The two lines below are Gemini generated: Prompt was that given a glm model, create automatic stepwise param selection
# Perform stepwise selection using AIC
# direction = "both" checks for variables to add AND remove at each step
final_model_daysmenthlthnotgood <- step(model_daysmenthlthnotgood, direction = "both")

# View the final selected model
summary(final_model_daysmenthlthnotgood)
tab_model(final_model_daysmenthlthnotgoodx, show.stat = FALSE)
```