---
title: "CDC BRFSS Analysis: Veteran Health"

author: "Math 62 - Spring 2026"
format:
  pdf:
    knitr:
      opts_chunk: 
        out-width: "60%"
        fig-align: "center"
        size: "scriptsize"
    fontsize: 10pt
    toc: false
    toc-params: "allowframebreaks"
    include-in-header:
      - text: |
          \usepackage{etoolbox}
          \usepackage{hyperref}
          
execute:
  echo: false
  warning: false
  message: false
  eval: true

fig-align: center
fig-height: 6
fig-width: 8

---

# Team Members

Li, Elsa

Louie, Amanda

Sharma, Anika

---

```{r}
#| echo: true
2 * 2
```

The `echo: true` option enables the printing of code (only output is displayed).


# EDA Section



```{r}
#| label: fig-status-dist
#| fig-cap: "Comparison of sample sizes between Veterans and Non-Veterans."

ggplot(brfss %>% filter(!is.na(veteran)), aes(x = veteran)) +
  geom_bar(fill = vet_blue, width = 0.6) +
  geom_text(stat = "count", aes(label = scales::comma(after_stat(count))), 
            vjust = -0.5, size = 4, fontface = "bold") +
  labs(title = "Distribution of Veterans vs Non-Veterans",
       x = "Veteran Status",
       y = "Total Count") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.x = element_blank())

```



```{r}
#| label: fig-log-reg
#| fig-height: 7
#| fig-width: 10

# Prepare data
model_data <- brfss %>%
  filter(!is.na(veteran)) %>%
  mutate(
    veteran_bin = as.integer(veteran == "Yes"),
    across(where(is.character), as.factor)
  ) %>%
  select(-veteran, -state, -adults) %>% 
  na.omit()

# Fit model
log_model <- glm(veteran_bin ~ ., data = model_data, family = binomial)

# Extract and format coefficients
coef_df <- as.data.frame(summary(log_model)$coefficients) %>%
  mutate(term = rownames(.)) %>%
  rename(estimate = Estimate, std_error = `Std. Error`, p_value = `Pr(>|z|)`) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    odds_ratio = exp(estimate),
    ci_low  = exp(estimate - 1.96 * std_error),
    ci_high = exp(estimate + 1.96 * std_error),
    significant = p_value < 0.05
  ) %>%
  arrange(desc(abs(estimate))) %>%
  slice_head(n = 20)

# Plot using the custom color scheme
ggplot(coef_df, aes(x = odds_ratio, y = reorder(term, odds_ratio), color = significant)) +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = ci_low, xmax = ci_high), height = 0.3) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "gray40") +
  scale_color_manual(values = c("FALSE" = "gray70", "TRUE" = vet_blue),
                     labels = c("Not significant", "p < 0.05")) +
  labs(
    title = "Top 20 Predictors of Veteran Status",
    subtitle = "Odds ratios calculated via logistic regression",
    x = "Odds Ratio (95% CI)",
    y = NULL,
    color = "Significance"
  ) +
  theme_minimal()
```


```{r}
#| label: fig-demographics
#| fig-height: 10
#| fig-width: 12
#| label: fig-vet-gender
#| fig-cap: "Gender distribution among surveyed veterans."

ggplot(vets, aes(x = sex)) +
  geom_bar(fill = "#29b9d6", width = 0.5) +
  geom_text(stat = "count", aes(label = scales::comma(after_stat(count))), 
            vjust = -0.5, size = 3.5, fontface = "bold") +
  labs(title = "Veteran Population by Gender", x = NULL, y = "Count") +
  theme_minimal()
```




```{r}
#| label: fig-vet-age
#| fig-cap: "Age distribution of veterans, showing a heavy skew toward older demographics."

#| label: fig-vet-race
#| fig-cap: "Racial demographics of veterans within the dataset."

ggplot(vets, aes(y = reorder(race, race, function(x) length(x)))) +
  geom_bar(fill = "#2a79d5") +
  geom_text(stat = "count", aes(label = scales::comma(after_stat(count))), 
            hjust = -0.1, size = 3) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.2))) +
  labs(title = "Veteran Population by Race", x = "Count", y = NULL) +
  theme_minimal()

```

```{r}
#| label: fig-education-comparison
#| fig-cap: "Education level distribution by veteran status."

brfss %>%
  filter(!is.na(veteran), !is.na(education)) %>%
  count(veteran, education) %>%
  group_by(veteran) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup() %>%
  ggplot(aes(x = education, y = prop, fill = veteran)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_text(
    aes(label = scales::percent(prop, accuracy = 0.1)),
    position = position_dodge(width = 0.7),
    vjust = -0.3,
    size = 3
  ) +
  scale_y_continuous(
    labels = scales::percent,
    expand = expansion(mult = c(0, 0.15))
  ) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 12)) +
  scale_fill_manual(values = c("No" = vet_teal, "Yes" = vet_blue)) +
  labs(
    title = "Education Distribution by Veteran Status",
    x = "Education Level",
    y = "Percent Within Group",
    fill = "Veteran"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```





```{r}
#| label: fig-income-comparison
#| fig-cap: "Income level distribution by veteran status."

brfss %>%
  filter(!is.na(veteran), !is.na(income)) %>%
  count(veteran, income) %>%
  group_by(veteran) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup() %>%
  ggplot(aes(x = income, y = prop, fill = veteran)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_text(
    aes(label = scales::percent(prop, accuracy = 0.1)),
    position = position_dodge(width = 0.7),
    vjust = -0.3,
    size = 3
  ) +
  scale_y_continuous(
    labels = scales::percent,
    expand = expansion(mult = c(0, 0.15))
  ) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 12)) +
  scale_fill_manual(values = c("No" = vet_teal, "Yes" = vet_blue)) +
  labs(
    title = "Income Distribution by Veteran Status",
    x = "Income Level",
    y = "Percent Within Group",
    fill = "Veteran"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```


```{r}
#| label: fig-age-comparison
#| fig-cap: "Age distribution by veteran status."

brfss %>%
  filter(!is.na(veteran), !is.na(age)) %>%
  count(veteran, age) %>%
  group_by(veteran) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup() %>%
  ggplot(aes(x = age, y = prop, fill = veteran)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_text(
    aes(label = scales::percent(prop, accuracy = 0.1)),
    position = position_dodge(width = 0.7),
    vjust = -0.3,
    size = 3
  ) +
  scale_y_continuous(
    labels = scales::percent,
    expand = expansion(mult = c(0, 0.15))
  ) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 12)) +
  scale_fill_manual(values = c("No" = vet_teal, "Yes" = vet_blue)) +
  labs(
    title = "Age Distribution by Veteran Status",
    x = "Age Group",
    y = "Percent Within Group",
    fill = "Veteran"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```


# Demographics Section


```{r}
library(dplyr)
library(tidyverse)
```
```{r}

# load in BRFSS dataset:

brfss = read.csv("brfss2022_m62.csv", stringsAsFactors=TRUE)


```



```{r}

# Running Chisq test to compare proportions between veterans and national data (gemini-gen)
# 1. Clean the data: Filter for Veterans and non-NA race entries
brfss_veterans <- brfss %>%
  filter(veteran == "Yes") %>%
  filter(!is.na(race))

# 2. Count the number of unique entries in the race column for veterans
vet_race_counts <- table(brfss_veterans$race)

# 3. Expected national probabilities (must sum to 1 and match table order)
expected_probs <- c(0.007, 0.061, 0.126, 0.191, 0.024, 0.002, 0.589)

# 4. Run the Chi-Squared Goodness of Fit Test
chisq.test(vet_race_counts, p = expected_probs)


```
```{r} 
# 1. Reshape the data for plotting
plot_data <- final_display %>%
  tidyr::pivot_longer(
    cols = c(Expected_Count, Observed_Count), 
    names_to = "Count_Type", 
    values_to = "Count_Value"
  )

# 2. Create the Grouped Bar Plot with Log Scale
ggplot(plot_data, aes(x = Race, y = Count_Value, fill = Count_Type)) +
  geom_bar(stat = "identity", position = "dodge") +
  # Apply the log transformation
  # 'labels = scales::comma' makes the numbers easier to read (e.g., 1,000 instead of 1e3)
#  scale_y_log10(labels = scales::comma) + 
  scale_x_discrete(labels = function(x) str_wrap(x, width = 15)) +
  theme_minimal(base_size=14) +
  labs(
    title = "Observed vs. Expected Veteran Counts by Race (Log Scale)",
    x = "Racial Group",
    y = "Count (Log 10)",
    fill = "Legend"
  ) +
  scale_fill_manual(
    values = c("Expected_Count" = "#2a79d5", "Observed_Count" = "#29b9d6"),
    labels = c("Expected", "Observed")
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor.y = element_blank() # Clean up log grid lines
  )
```


```{r}
# Compare veteran race proportions against national expected proportions

observed_counts <- as.numeric(vet_race_counts)
race_names <- names(vet_race_counts)
n_total <- sum(observed_counts)

race_prop_tests <- data.frame(
  Race = race_names,
  Observed_Count = observed_counts,
  Observed_Prop = observed_counts / n_total,
  Expected_Prop = expected_probs
) %>%
  rowwise() %>%
  mutate(
    p_value = prop.test(
      x = Observed_Count,
      n = n_total,
      p = Expected_Prop,
      correct = FALSE
    )$p.value
  ) %>%
  ungroup() %>%
  mutate(
    p_adj_bonferroni = p.adjust(p_value, method = "bonferroni"),
    difference = Observed_Prop - Expected_Prop,
    abs_difference = abs(difference)
  ) %>%
  arrange(p_value)

race_prop_tests
```


```{r}
race_prop_tests %>%
  mutate(direction = ifelse(difference > 0, "Overrepresented", "Underrepresented")) %>%
  select(Race, Observed_Prop, Expected_Prop, direction, p_adj_bonferroni)
```






---
title: "Veteran Demographics Analysis"
format: html
---

```{r}
#| label: setup
#| message: false
#| warning: false

library(dplyr)
library(gt)

# 1. Recreating your processed dataframe based on the screenshot
# (In your actual file, you can just use 'race_prop_tests' from your previous code)
race_prop_tests <- data.frame(
  Race = c("Asian only, non-Hispanic", "Hispanic", "White only, non-Hispanic", 
           "Black only, non-Hispanic", "American Indian or Alaskan Native only, Non-Hispanic",
           "Native Hawaiian or other Pacific Islander only, Non-Hispanic", "Multiracial, non-Hispanic"),
  Observed_Prop = c(0.012736694, 0.049953993, 0.813162865, 0.074313526, 0.017119473, 0.006344133, 0.026369316),
  Expected_Prop = c(0.061, 0.191, 0.589, 0.126, 0.007, 0.002, 0.024),
  p_adj_bonferroni = c(0, 0, 0, 4.990780e-219, 1.735058e-133, 4.604660e-86, 1.158692e-02)
) %>%
  mutate(direction = ifelse(Observed_Prop > Expected_Prop, "Overrepresented", "Underrepresented"))

# 2. Generate the formatted table
race_prop_tests %>%
  gt() %>%
  # Relabel columns to match your desired style
  cols_label(
    Race = "Race",
    Observed_Prop = "Veteran Prop.",
    Expected_Prop = "National Prop.",
    direction = "Representation",
    p_adj_bonferroni = "p-value (adj)"
  ) %>%
  # Format proportions as decimals (matching your first screenshot style)
  fmt_number(
    columns = c(Observed_Prop, Expected_Prop),
    decimals = 4
  ) %>%
  # Format p-values: small values as < 2.2e-16 or scientific notation
  fmt_scientific(
    columns = p_adj_bonferroni,
    decimals = 2
  ) %>%
  # Apply the "Academic/Minimalist" styling (Thin lines top and bottom)
  tab_options(
    table.border.top.color = "black",
    table.border.top.width = px(2),
    table.border.bottom.color = "black",
    table.border.bottom.width = px(2),
    column_labels.border.bottom.color = "black",
    column_labels.border.bottom.width = px(1),
    table_body.hlines.style = "none", # Removes horizontal lines between rows
    data_row.padding = px(4)
  ) %>%
  # Align text left and numbers right
  cols_align(
    align = "left",
    columns = Race
  ) %>%
  # Handle the extremely small p-values to look like the screenshot
  text_transform(
    locations = cells_body(columns = p_adj_bonferroni, rows = p_adj_bonferroni < 2.2e-16),
    fn = function(x) "< 2.2e-16"
  )



```




```{r}

# measuring differences in Demographics: education level

brfss_veterans_edu <- brfss %>%
  filter(veteran == "Yes", age >= 25, !is.na(education))

table(brfss$education, useNA = "ifany")


brfss_veterans_edu <- brfss %>%
  filter(veteran == "Yes", age >= 25, !is.na(education)) %>%
  mutate(edu_group = case_when(
    education %in% c("No school", "Elementary", "Some HS") ~ "Less than high school",
    education == "HS Grad" ~ "High school graduate",
    education == "Some College" ~ "Some college",
    education == "College Grad" ~ "Bachelor's or higher",
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(edu_group))
```


```{r}
# Education comparison: BRFSS veterans vs 2022 U.S. national education distribution
## Education comparison: BRFSS veterans vs 2022 U.S. national education distribution
# Source: U.S. Census ACS 2022 S1501, population age 25+

library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)
library(readr)

# Census ACS 2022 proportions, age 25+
# These are rounded published values, so we normalize them below.
expected_edu_props <- c(
  "Less than high school" = 0.109,
  "High school graduate" = 0.264,
  "Some college / associate" = 0.285,
  "Bachelor's or higher" = 0.209 + 0.134
)

# Force probabilities to sum exactly to 1
expected_edu_props <- expected_edu_props / sum(expected_edu_props)

# Clean BRFSS veteran education data
brfss_veterans_edu <- brfss %>%
  filter(veteran == "Yes", !is.na(education)) %>%
  mutate(
    age_num = if ("age" %in% names(.)) parse_number(as.character(age)) else NA_real_,
    edu_group = case_when(
      education %in% c("No school", "Elementary", "Some HS") ~ "Less than high school",
      education == "HS Grad" ~ "High school graduate",
      education == "Some College" ~ "Some college / associate",
      education == "College Grad" ~ "Bachelor's or higher",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(edu_group)) %>%
  filter(is.na(age_num) | age_num >= 25)

# Observed counts
observed_edu_counts <- table(brfss_veterans_edu$edu_group)

# Reorder expected probabilities to match observed table
expected_edu_props_ordered <- expected_edu_props[names(observed_edu_counts)]

# Normalize again after reordering, just in case any categories are missing
expected_edu_props_ordered <- expected_edu_props_ordered / sum(expected_edu_props_ordered)

# Chi-square goodness-of-fit test
edu_chisq_test <- chisq.test(
  observed_edu_counts,
  p = expected_edu_props_ordered
)

edu_chisq_test

# Table showing which education groups differ most
edu_diff_table <- data.frame(
  Education = names(observed_edu_counts),
  Observed_Count = as.numeric(observed_edu_counts),
  Expected_Prop = expected_edu_props_ordered
) %>%
  mutate(
    Total = sum(Observed_Count),
    Observed_Prop = Observed_Count / Total,
    Expected_Count = Expected_Prop * Total,
#    Difference = Observed_Prop - Expected_Prop,
#    Percent_Point_Gap = 100 * Difference,
 #   ChiSq_Contribution = (Observed_Count - Expected_Count)^2 / Expected_Count,
    Direction = ifelse(Difference > 0, "Overrepresented", "Underrepresented")
  ) %>%
  arrange(desc(ChiSq_Contribution))

edu_diff_table

# Education group that differs most
edu_diff_table %>%
  slice_max(ChiSq_Contribution, n = 1)

# Plot observed vs expected proportions
edu_diff_table %>%
  select(Education, Observed_Prop, Expected_Prop) %>%
  pivot_longer(
    cols = c(Observed_Prop, Expected_Prop),
    names_to = "Type",
    values_to = "Proportion"
  ) %>%
  ggplot(aes(x = Education, y = Proportion, fill = Type)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Veteran Education Distribution vs 2022 U.S. National Average",
    x = "Education Level",
    y = "Proportion",
    fill = "Group"
  ) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```


```{r}
# One-sample proportion tests for each education group
# with Bonferroni correction

edu_prop_tests <- data.frame(
  Education = names(observed_edu_counts),
  Observed_Count = as.numeric(observed_edu_counts),
  Expected_Prop = expected_edu_props_ordered
) %>%
  mutate(
    Total = sum(Observed_Count),
    Observed_Prop = Observed_Count / Total,
    Difference = Observed_Prop - Expected_Prop,
  ) %>%
  rowwise() %>%
  mutate(
    p_value = prop.test(
      x = Observed_Count,
      n = Total,
      p = Expected_Prop,
      correct = FALSE
    )$p.value
  ) %>%
  ungroup() %>%
  mutate(
    p_bonferroni = p.adjust(p_value, method = "bonferroni"),
    Significant_Bonferroni = p_bonferroni < 0.05,
    Direction = ifelse(Difference > 0, "Overrepresented", "Underrepresented")
  ) %>%
  arrange(p_bonferroni)

edu_prop_tests <- edu_prop_tests %>%
  select(
    Education,
    Observed_Prop,
    Expected_Prop,
    p_bonferroni,
    Direction
  )

edu_prop_tests
```




```{r}
# Income comparison: BRFSS veterans vs 2022 U.S. national household income distribution

library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)

# Manually entered 2022 ACS national household income proportions
# Broad categories based on Census ACS B19001
national_income_df <- data.frame(
  Income = c(
    "Less than $25,000",
    "$25,000 to less than $50,000",
    "$50,000 to less than $75,000",
    "$75,000 or more"
  ),
  Expected_Prop = c(
    0.161,
    0.171,
    0.148,
    0.520
  )
) %>%
  mutate(Expected_Prop = Expected_Prop / sum(Expected_Prop))

# Clean BRFSS veteran income data
brfss_veterans_income <- brfss %>%
  filter(veteran == "Yes", !is.na(income)) %>%
  mutate(
    income_chr = as.character(income),
    income_group = case_when(
      income_chr %in% c(
        "Less than $10000",
        "$10000 to $14999",
        "$15000 to $19999",
        "$20000 to $24999"
      ) ~ "Less than $25,000",
      
      income_chr %in% c(
        "$25000 to $34999",
        "$35000 to $49999"
      ) ~ "$25,000 to less than $50,000",
      
      income_chr == "$50000 to $74999" ~ "$50,000 to less than $75,000",
      
      income_chr %in% c(
        "$75000 to $99999",
        "$100000 to $149999",
        "$150000 to $199999",
        "$200000 or more"
      ) ~ "$75,000 or more",
      
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(income_group))

# Observed veteran income counts
observed_income_counts <- table(brfss_veterans_income$income_group)

# Expected proportions ordered to match observed table
expected_income_props_ordered <- national_income_df$Expected_Prop[
  match(names(observed_income_counts), national_income_df$Income)
]

expected_income_props_ordered <- expected_income_props_ordered / sum(expected_income_props_ordered)

# Chi-square goodness-of-fit test
income_chisq_test <- chisq.test(
  observed_income_counts,
  p = expected_income_props_ordered
)

income_chisq_test

# Table showing which income groups differ most
income_diff_table <- data.frame(
  Income = names(observed_income_counts),
  Observed_Count = as.numeric(observed_income_counts),
  Expected_Prop = expected_income_props_ordered
) %>%
  mutate(
    Total = sum(Observed_Count),
    Observed_Prop = Observed_Count / Total,
    Expected_Count = Expected_Prop * Total,
    Difference = Observed_Prop - Expected_Prop,
    ChiSq_Contribution = (Observed_Count - Expected_Count)^2 / Expected_Count,
    Direction = ifelse(Difference > 0, "Overrepresented", "Underrepresented")
  ) %>%
  arrange(desc(ChiSq_Contribution))

income_diff_table

# One-sample proportion tests with Bonferroni correction
income_prop_tests <- data.frame(
  Income = names(observed_income_counts),
  Observed_Count = as.numeric(observed_income_counts),
  Expected_Prop = expected_income_props_ordered
) %>%
  mutate(
    Total = sum(Observed_Count),
    Observed_Prop = Observed_Count / Total,
    Difference = Observed_Prop - Expected_Prop
  ) %>%
  rowwise() %>%
  mutate(
    p_value = prop.test(
      x = Observed_Count,
      n = Total,
      p = Expected_Prop,
      correct = FALSE
    )$p.value
  ) %>%
  ungroup() %>%
  mutate(
    p_bonferroni = p.adjust(p_value, method = "bonferroni"),
    Direction = ifelse(Difference > 0, "Overrepresented", "Underrepresented")
  ) %>%
  select(
    Income,
    Observed_Prop,
    Expected_Prop,
    p_bonferroni,
    Direction
  ) %>%
  arrange(p_bonferroni)

income_prop_tests
# Plot observed vs expected proportions using same color scheme
income_diff_table %>%
  select(Income, Observed_Prop, Expected_Prop) %>%
  pivot_longer(
    cols = c(Observed_Prop, Expected_Prop),
    names_to = "Type",
    values_to = "Proportion"
  ) %>%
  mutate(
    Type = recode(
      Type,
      "Observed_Prop" = "Observed",
      "Expected_Prop" = "Expected"
    )
  ) %>%
  ggplot(aes(x = Income, y = Proportion, fill = Type)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(
    values = c("Expected" = vet_blue, "Observed" = vet_teal)
  ) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Veteran Income Distribution vs 2022 U.S. National Average",
    x = "Income Level",
    y = "Proportion",
    fill = "Group"
  ) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```















# Veteran vs Non Veteran: Mental + Physical Health

#Amanda Section 

```{r}
#| echo: false
library(dplyr)
brfss = read.csv("brfss2022_m62.csv", stringsAsFactors=TRUE)
veteran <- subset(brfss, veteran == "Yes")
nrow(veteran)
nonVeteran <- subset(brfss, veteran == "No")
nrow(nonVeteran)
```

Comparisons of physical health between veterans and non-veterans:

Arthritis: p-value $< 2.2 \times 10^{-16}$
```{r}
#| echo: false

veteranArthClean <- veteran %>% filter(!is.na(arthritis))
nonVeteranArthClean <- nonVeteran %>% filter(!is.na(arthritis))

arthritisVeterans <- sum(veteranArthClean$arthritis == "Yes")
arthritisNonVeterans <- sum(nonVeteranArthClean$arthritis == "Yes")

successes <- c(arthritisVeterans, arthritisNonVeterans)
n_obs <- c(nrow(veteranArthClean), nrow(nonVeteranArthClean))

arthritis_prop_test <- prop.test(successes, n_obs)
arthritis_prop_test
```

Heart Disease: p-value $< 2.2 \times 10^{-16}$
```{r}
#| echo: false

veteranHDClean <- veteran %>% filter(!is.na(heartdisease))
nonVeteranHDClean <- nonVeteran %>% filter(!is.na(heartdisease))

hdVeterans <- sum(veteranHDClean$heartdisease == "Yes")
hdNonVeterans <- sum(nonVeteranHDClean$heartdisease == "Yes")

successes <- c(hdVeterans, hdNonVeterans)
n_obs <- c(nrow(veteranHDClean), nrow(nonVeteranHDClean))

hd_prop_test <- prop.test(successes, n_obs)
hd_prop_test
```

Cancer: p-value $< 2.2 \times 10^{-16}$
```{r}
#| echo: false

veteranCancerClean <- veteran %>% filter(!is.na(cancer))
nonVeteranCancerClean <- nonVeteran %>% filter(!is.na(cancer))

cancerVeterans <- sum(veteranCancerClean$cancer == "Yes")
cancerNonVeterans <- sum(nonVeteranCancerClean$cancer == "Yes")

successes <- c(cancerVeterans, cancerNonVeterans)
n_obs <- c(nrow(veteranCancerClean), nrow(nonVeteranCancerClean))

cancer_prop_test <- prop.test(successes, n_obs)
cancer_prop_test
```

Diabetes: p-value $< 2.2 \times 10^{-16}$
```{r}
#| echo: false

veteranDiabetesClean <- veteran %>% filter(!is.na(diabetes))
nonVeteranDiabetesClean <- nonVeteran %>% filter(!is.na(diabetes))

diabetesVeterans <- sum(veteranDiabetesClean$diabetes == "Yes")
diabetesNonVeterans <- sum(nonVeteranDiabetesClean$diabetes == "Yes")

successes <- c(diabetesVeterans, diabetesNonVeterans)
n_obs <- c(nrow(veteranDiabetesClean), nrow(nonVeteranDiabetesClean))

diabetes_prop_test <- prop.test(successes, n_obs)
diabetes_prop_test
```

COPD: p-value $< 2.2 \times 10^{-16}$
```{r}
#| echo: false

veteranCOPDClean <- veteran %>% filter(!is.na(copd))
nonVeteranCOPDClean <- nonVeteran %>% filter(!is.na(copd))

copdVeterans <- sum(veteranCOPDClean$copd == "Yes")
copdNonVeterans <- sum(nonVeteranCOPDClean$copd == "Yes")

successes <- c(copdVeterans, copdNonVeterans)
n_obs <- c(nrow(veteranCOPDClean), nrow(nonVeteranCOPDClean))

copd_prop_test <- prop.test(successes, n_obs)
copd_prop_test
```

daysphyshlthnotgood: p-value $< 2.2 \times 10^{-16}$
```{r}
#| echo: false

veteranDaysphyshlthnotgoodClean <- veteran %>% filter(!is.na(daysphyshlthnotgood))
nonVeteranDaysphyshlthnotgoodClean <- nonVeteran %>% filter(!is.na(daysphyshlthnotgood))

daysphyshlthnotgoodVeterans <- mean(veteranDaysphyshlthnotgoodClean$daysphyshlthnotgood)
daysphyshlthnotgoodNonVeterans <- mean(nonVeteranDaysphyshlthnotgoodClean$daysphyshlthnotgood)

daysphyshlthnotgood_t_test <- t.test(veteranDaysphyshlthnotgoodClean$daysphyshlthnotgood,
                                     nonVeteranDaysphyshlthnotgoodClean$daysphyshlthnotgood,
                                     alternative="two.sided",
                                     mu=0, paired=FALSE, var.equal = FALSE,
                                     conf.level=0.95)
daysphyshlthnotgood_t_test
```

Check Assumptions (equal variances):
```{r}
#| echo: false
#| 
hist(veteranDaysphyshlthnotgoodClean$daysphyshlthnotgood,
     main = "Days Physical Health Not Good (Veteran)",
     xlab = "Number of days in past month physical health was not good")
hist(nonVeteranDaysphyshlthnotgoodClean$daysphyshlthnotgood,
     main = "Days Physical Health Not Good (Non-Veteran)",
     xlab = "Number of days in past month physical health was not good")
```


```{r}
#| echo: false
# Summary of physical health tests

factors <- c("Arthritis", "Heart Disease", "Cancer", "Diabetes", "COPD", "Days Physical Health Not Good")
prop_vet <- c(0.4356422, 0.1244373, 0.1789827, 0.1962027, 0.11969946, 4.972634)
prop_nonVet <- c(0.3455287, 0.0543102, 0.1105903, 0.1340290, 0.07724301, 4.338173)
p_values <- rep("< 2.2e-16", 6)

physical_tests <- data.frame(
  Factors = factors,
  Veteran = prop_vet,
  NonVeteran = prop_nonVet,
  pvalue = p_values
)

physical_tests
```

```{r}
#| echo: false
# generated with help from gemini

# Your vectors
labels_a <- c("Arthritis", "Heart Disease", "Cancer", "Diabetes", "COPD")
prop_vet <- c(0.4356422, 0.1244373, 0.1789827, 0.1962027, 0.11969946)
prop_nonVet <- c(0.3455287, 0.0543102, 0.1105903, 0.1340290, 0.07724301)

# Combine into a matrix
data_matrix_a <- rbind(prop_vet, prop_nonVet)

# Assign column names to the matrix
colnames(data_matrix_a) <- labels_a

# Your vectors
labels_b <- c("daysphyshlthnotgood")
mean_vet <- c(4.972634)
mean_nonVet <- c(4.338173)

# Combine into a matrix
data_matrix_b <- rbind(mean_vet, mean_nonVet)

# Assign column names to the matrix
colnames(data_matrix_b) <- labels_b

# Create the grouped barplot
# Adjust margins so the long label fits (bottom, left, top, right)
par(mar = c(12, 4, 4, 2) + 0.1)
par(mfrow=c(1,2))

barplot(data_matrix_a, 
        beside = TRUE,                   # This places the bars side-by-side
        col = c("skyblue", "salmon"),    # Colors for the two groups
        legend.text = c("Veteran", "Non-Veteran"),
        args.legend = list(x = "topright", bty = "n"),
        main = "Health Indicators: ",
        ylab = "Value",
        las = 1,                         # Rotates x-axis labels to be vertical
        cex.names = 0.8)                 # Shrinks the label text slightly

barplot(data_matrix_b, 
        beside = TRUE,                   # This places the bars side-by-side
        col = c("skyblue", "salmon"),    # Colors for the two groups
        legend.text = c("Veteran", "Non-Veteran"),
        args.legend = list(x = "topright", bty = "n"),
        main = "Veterans vs Non-Veterans",
        ylab = "Value",
        las = 1,                         # Rotates x-axis labels to be vertical
        cex.names = 0.8)                 # Shrinks the label text slightly
```

Comparisons of mental health between veterans and non-veterans:

Depression: p-value $< 2.2 \times 10^{-16}$
```{r}
#| echo: false

veteranDepressClean <- veteran %>% filter(!is.na(depress))
nonVeteranDepressClean <- nonVeteran %>% filter(!is.na(depress))

depressVeterans <- sum(veteranDepressClean$depress == "Yes")
depressNonVeterans <- sum(nonVeteranDepressClean$depress == "Yes")

successes <- c(depressVeterans, depressNonVeterans)
n_obs <- c(nrow(veteranDepressClean), nrow(nonVeteranDepressClean))

depress_prop_test <- prop.test(successes, n_obs)
depress_prop_test
```

daysmenthlthnotgood: p-value $< 2.2 \times 10^{-16}$
```{r}
#| echo: false

veteranDaysmenthlthnotgoodClean <- veteran %>% filter(!is.na(daysmenthlthnotgood))
nonVeteranDaysmenthlthnotgoodClean <- nonVeteran %>% filter(!is.na(daysmenthlthnotgood))

daysmenthlthnotgoodVeterans <- mean(veteranDaysmenthlthnotgoodClean$daysmenthlthnotgood)
daysmenthlthnotgoodNonVeterans <- mean(nonVeteranDaysmenthlthnotgoodClean$daysmenthlthnotgood)

daysmenthlthnotgood_t_test <- t.test(veteranDaysmenthlthnotgoodClean$daysmenthlthnotgood,
                                     nonVeteranDaysmenthlthnotgoodClean$daysmenthlthnotgood,
                                     alternative="two.sided",
                                     mu=0, paired=FALSE, var.equal = TRUE,
                                     conf.level=0.95)
daysmenthlthnotgood_t_test
```

Check Assumptions (equal variances):
```{r}
#| echo: false
#| 
hist(veteranDaysmenthlthnotgoodClean$daysmenthlthnotgood,
     main = "Days Mental Health Not Good (Veteran)",
     xlab = "Number of days in past month mental health was not good")
hist(nonVeteranDaysmenthlthnotgoodClean$daysmenthlthnotgood,
     main = "Days Mental Health Not Good (Non-Veteran)",
     xlab = "Number of days in past month mental health was not good")
```

```{r}
#| echo: false
# Summary of Mental Health Tests

factors <- c("Depression", "Days Physical Health Not Good")
prop_vet <- c(0.1757344, 3.517690)
prop_nonVet <- c(0.2166535, 4.462462)
p_values <- rep("< 2.2e-16", 2)

mental_tests <- data.frame(
  Factors = factors,
  Veteran = prop_vet,
  NonVeteran = prop_nonVet,
  pvalue = p_values
)

mental_tests
```

```{r}
#| echo: false
# generated with help from gemini

# Your vectors
labels_a <- c("Depression")
prop_vet <- c(0.1757344)
prop_nonVet <- c(0.2166535)

# Combine into a matrix
data_matrix_a <- rbind(prop_vet, prop_nonVet)

# Assign column names to the matrix
colnames(data_matrix_a) <- labels_a

# Your vectors
labels_b <- c("daysmenthlthnotgood")
mean_vet <- c(3.517690)
mean_nonVet <- c(4.462462)

# Combine into a matrix
data_matrix_b <- rbind(mean_vet, mean_nonVet)

# Assign column names to the matrix
colnames(data_matrix_b) <- labels_b

# Create the grouped barplot
# Adjust margins so the long label fits (bottom, left, top, right)
par(mar = c(12, 4, 4, 2) + 0.1)
par(mfrow=c(1,2))

barplot(data_matrix_a, 
        beside = TRUE,                   # This places the bars side-by-side
        col = c("skyblue", "salmon"),    # Colors for the two groups
        legend.text = c("Veteran", "Non-Veteran"),
        args.legend = list(x = "topright", bty = "n"),
        main = "Health Indicators: ",
        ylab = "Value",
        las = 1,                         # Rotates x-axis labels to be vertical
        cex.names = 0.8)                 # Shrinks the label text slightly

barplot(data_matrix_b, 
        beside = TRUE,                   # This places the bars side-by-side
        col = c("skyblue", "salmon"),    # Colors for the two groups
        legend.text = c("Veteran", "Non-Veteran"),
        args.legend = list(x = "topright", bty = "n"),
        main = "Veterans vs Non-Veterans",
        ylab = "Value",
        las = 1,                         # Rotates x-axis labels to be vertical
        cex.names = 0.8)                 # Shrinks the label text slightly
```

What is the difference in the probability of having a specific health factor being a veteran vs. non-veteran?

```{r}
#| echo: false

arthritisClean <- brfss %>% filter(!is.na(arthritis))

arthritisClean$veteran <- as.factor(arthritisClean$veteran)

arthritisClean$arthritis <- as.factor(arthritisClean$arthritis)

arthritisMod <- glm(formula = arthritis ~ veteran, data = arthritisClean, family = "binomial")

summary(arthritisMod)
```

```{r}
p_vet_arthritis <- plogis(-0.63875 + 0.37988 * 1)
p_nonVet_arthritis <- plogis(-0.63875 + 0.37988 * 0)

cat("P(Arthritis | Veteran) = ", p_vet_arthritis)
cat('\n')
cat("P(Arthritis | Non-Veteran) = ", p_nonVet_arthritis)
```

```{r}
#| echo: false

hdClean <- brfss %>% filter(!is.na(heartdisease))

hdClean$veteran <- as.factor(hdClean$veteran)

hdClean$heartdisease <- as.factor(hdClean$heartdisease)

hdMod <- glm(formula = heartdisease ~ veteran, data = hdClean, family = "binomial")

summary(hdMod)
```

```{r}
p_vet_hd <- plogis(-2.857202 + 0.906138 * 1)
p_nonVet_hd <- plogis(-2.857202 + 0.906138 * 0)

cat("P(Heart Disease | Veteran) = ", p_vet_hd)
cat('\n')
cat("P(Heart Disease | Non-Veteran) = ", p_nonVet_hd)
```

```{r}
#| echo: false

cancerClean <- brfss %>% filter(!is.na(cancer))

cancerClean$veteran <- as.factor(cancerClean$veteran)

cancerClean$cancer <- as.factor(cancerClean$cancer)

cancerMod <- glm(formula = cancer ~ veteran, data = cancerClean, family = "binomial")

summary(cancerMod)
```

```{r}
p_vet_cancer <- plogis(-2.084725 + 0.56147 * 1)
p_nonVet_cancer <- plogis(-2.084725 + 0.56147 * 0)

cat("P(Cancer | Veteran) = ", p_vet_cancer)
cat('\n')
cat("P(Cancer | Non-Veteran) = ", p_nonVet_cancer)
```

```{r}
#| echo: false

diabetesClean <- brfss %>% filter(!is.na(diabetes))

diabetesClean$veteran <- as.factor(diabetesClean$veteran)

diabetesClean$diabetes <- as.factor(diabetesClean$diabetes)

diabetesMod <- glm(formula = diabetes ~ veteran, data = diabetesClean, family = "binomial")

summary(diabetesMod)
```

```{r}
p_vet_diabetes <- plogis(-1.602633 + 0.370480 * 1)
p_nonVet_diabetes <- plogis(-1.602633 + 0.370480 * 0)

cat("P(Diabetes | Veteran) = ", p_vet_diabetes)
cat('\n')
cat("P(Diabetes | Non-Veteran) = ", p_nonVet_diabetes)
```

```{r}
#| echo: false

copdClean <- brfss %>% filter(!is.na(copd))

copdClean$veteran <- as.factor(copdClean$veteran)

copdClean$diabetes <- as.factor(copdClean$copd)

copdMod <- glm(formula = copd ~ veteran, data = copdClean, family = "binomial")

summary(copdMod)
```

```{r}
p_vet_copd <- plogis(-2.480409 + 0.485130 * 1)
p_nonVet_copd <- plogis(-2.480409 + 0.485130 * 0)

cat("P(COPD | Veteran) = ", p_vet_copd)
cat('\n')
cat("P(COPD | Non-Veteran) = ", p_nonVet_copd)
```

```{r}
#| echo: false

depressClean <- brfss %>% filter(!is.na(depress))

depressClean$veteran <- as.factor(depressClean$veteran)

depressClean$depress <- as.factor(depressClean$depress)

depressMod <- glm(formula = depress ~ veteran, data = depressClean, family = "binomial")

summary(depressMod)
```

```{r}
p_vet_depress <- plogis(-1.285276 - 0.260243 * 1)
p_nonVet_depress <- plogis(-1.285276 - 0.260243 * 0)

cat("P(Depressive Disorder Diagnosis | Veteran) = ", p_vet_depress)
cat('\n')
cat("P(Depressive Disorder Diagnosis | Non-Veteran) = ", p_nonVet_depress)
```

# Factors that impact veteran physical + mental health

#Elsa Section

What are some factors that influence physical health among veterans

```{r}
library(sjPlot)
library(sjlabelled)
library(sjmisc)
library(ggplot2)
brfss = read.csv("brfss2022_m62.csv", stringsAsFactors=TRUE)

head(brfss)
```

```{r Physical Health: Heart Disease}
#| echo: false
veteran <- subset(brfss, veteran == "Yes")


# Code below is Gemini generated: Prompt was that given these age groups, merge them so that we have three columns of young adult, middle age, and seniors

# 1. Create the new categorization logic
# We assign the categories based on the modified strings
veteran$age_category <- ifelse(veteran$age %in% c("25 to 29", "30 to 34", "35 to 39"), "young adult",
                        ifelse(veteran$age %in% c("40 to 44", "45 to 49", "50 to 54", "55 to 59", "60 to 64"), "middle-aged",
                        ifelse(veteran$age %in% c("65 to 69", "70 to 74", "75 to 79", "80 or older"), "senior", NA)))

# 2. Remove the old granular column and rename the new one to 'age'
veteran$age <- NULL
names(veteran)[names(veteran) == "age_category"] <- "age"

#Code below is Gemini generated: Prompt was that given these income groups, merge them so that we have three columns of high income, medium income, low income
# 1. Create the new categorization logic
veteran$income <- ifelse(veteran$income %in% c("$150000 to $199999", "$200000 or more"), 
                                "high income",
                         ifelse(veteran$income %in% c("Less than $10000", "$10000 to $14999", 
                                                             "$15000 to $19999", "$20000 to $24999", 
                                                             "$25000 to $34999"), 
                                "low income",
                         ifelse(is.na(veteran$income), NA, "medium income")))

# 2. Check the results
table(veteran$income, useNA = "always")

```


```{r}
# Two lines below generated by Gemini, prompt was that given these columns, drop any rows where any of these columns are NA
# Define the variables of interest
vars <- c("heartdisease", "sex", "bmi", "cancer", "copd", 
          "daysphyshlthnotgood", "anyexer", "race", 
          "age", "bingedrinker", "weeklydrinks", "heartdisease")

# Subset rows where all specified columns are non-NA
veteran_physical <- veteran[complete.cases(veteran[, vars]), ]
```


```{r}
#| echo: false
model_heartdisease <- glm(formula = heartdisease ~ sex + bmi + copd +
                 daysphyshlthnotgood + anyexer + race +
                 age + bingedrinker + weeklydrinks , data=veteran_physical, family=binomial)

```




```{r Physical Health: Cancer}
#| echo: false

model_cancer <- glm(formula = cancer ~ sex + bmi + copd +
                 daysphyshlthnotgood + anyexer + race +
                 age + bingedrinker + weeklydrinks , data=veteran_physical, family=binomial)

```


```{r Physical Health: Diabetes}
#| echo: false

model_diabetes <- glm(formula = diabetes ~ sex + bmi + copd +
                 daysphyshlthnotgood + anyexer + race +
                 age + bingedrinker + weeklydrinks , data=veteran_physical, family=binomial)

```

```{r Physical Health: BMI}
#| echo: false

model_bmi <- lm(formula = bmi ~ sex + copd +
                 daysphyshlthnotgood + anyexer + race +
                 age + bingedrinker + weeklydrinks , data=veteran_physical)

```

What are some factors that influence mental health among veterans

```{r Mental Health: Depression}
#| echo: false

# Two lines below generated by Gemini, prompt was that given these columns, drop any rows where any of these columns are NA
# Define the variables of interest

vars_mental <- c("daysmenthlthnotgood", "medcost", "income", "race", "age", "bingedrinker", "weeklydrinks", "sex", "depress")

# Subset rows where all specified columns are non-NA
veteran_mental <- veteran[complete.cases(veteran[, vars_mental]), ]


model_depression <- glm(formula = depress ~ sex + medcost + income + race +
                 age + bingedrinker + weeklydrinks + daysmenthlthnotgood , data=veteran_mental, family=binomial)

```


```{r Mental Health: Daysmenthlthnotgood}
#| echo: false

# Two lines below generated by Gemini, prompt was that given these columns, drop any rows where any of these columns are NA
# Define the variables of interest
model_daysmenthlthnotgood<- lm(formula = daysmenthlthnotgood ~ sex + medcost + income + race +
                 age + bingedrinker + weeklydrinks + depress , data=veteran_mental)

```

```{r}
#Below code generated by Gemini, instructions was to make a data frame table that included the below columns based on a given regression model

# 1. Updated Helper Function to include p-values
get_sig_df <- function(model) {
  # Extract coefficients into a data frame
  stats <- as.data.frame(summary(model)$coefficients)
  
  # Standardize column names
  colnames(stats) <- c("Coefficient", "StdError", "t_z_value", "p_value")
  
  # Move Predictors to a column
  stats$Predictor <- rownames(stats)
  
  # Calculate Odds Ratio
  stats$Odds_Ratio <- exp(stats$Coefficient)
  
  # Filter for significance (p < 0.01) to be on the safe side
  # Select Predictor, Coefficient, Odds_Ratio, and p_value
  final_df <- stats[stats$p_value < 0.01, c("Predictor", "Coefficient", "Odds_Ratio", "p_value")]
  
  # Reset row names
  rownames(final_df) <- NULL
  
  return(final_df)
}

# 2. Generate the 5 data frames
sig_heart    <- get_sig_df(model_heartdisease)
sig_cancer   <- get_sig_df(model_cancer)
sig_diabetes <- get_sig_df(model_diabetes)
sig_bmi      <- get_sig_df(model_bmi)
sig_depress  <- get_sig_df(model_depression)

# 3. Display the tables
cat("\n--- Table 1: Heart Disease ---\n")
print(sig_heart)

cat("\n--- Table 2: Cancer ---\n")
print(sig_cancer)

cat("\n--- Table 3: Diabetes ---\n")
print(sig_diabetes)

cat("\n--- Table 4: BMI ---\n")
print(sig_bmi)

cat("\n--- Table 5: Depression ---\n")
print(sig_depress)

```


```{r}
plot(model_bmi, which = 1)
plot(model_bmi, which = 2)
```