---
title: "CDC BRFSS Analysis: Veteran Health"

author: "Math 62 - Spring 2026"
format:
  pdf:
    knitr:
      opts_chunk: 
        out-width: "60%"
        fig-align: "center"
        size: "scriptsize"
    fontsize: 10pt
    toc: false
    toc-params: "allowframebreaks"
    include-in-header:
      - text: |
          \usepackage{etoolbox}
          \usepackage{hyperref}
          
execute:
  echo: false
  warning: false
  message: false
  eval: true

fig-align: center
fig-height: 6
fig-width: 8

---

# Team Members

Li, Elsa

Louie, Amanda

Sharma, Anika

---

```{r}
#| echo: true
2 * 2
```

The `echo: true` option enables the printing of code (only output is displayed).

Amanda Section 

```{r}
#| echo: false
library(dplyr)
brfss = read.csv("brfss2022_m62.csv", stringsAsFactors=TRUE)
veteran <- subset(brfss, veteran == "Yes")
nrow(veteran)
nonVeteran <- subset(brfss, veteran == "No")
nrow(nonVeteran)
```

Comparisons of physical health between veterans and non-veterans:

Arthritis: p-value $< 2.2 \times 10^{-16}$
```{r}
#| echo: false

veteranArthClean <- veteran %>% filter(!is.na(arthritis))
nonVeteranArthClean <- nonVeteran %>% filter(!is.na(arthritis))

arthritisVeterans <- sum(veteranArthClean$arthritis == "Yes")
arthritisNonVeterans <- sum(nonVeteranArthClean$arthritis == "Yes")

successes <- c(arthritisVeterans, arthritisNonVeterans)
n_obs <- c(nrow(veteranArthClean), nrow(nonVeteranArthClean))

arthritis_prop_test <- prop.test(successes, n_obs)
arthritis_prop_test
```

Heart Disease: p-value $< 2.2 \times 10^{-16}$
```{r}
#| echo: false

veteranHDClean <- veteran %>% filter(!is.na(heartdisease))
nonVeteranHDClean <- nonVeteran %>% filter(!is.na(heartdisease))

hdVeterans <- sum(veteranHDClean$heartdisease == "Yes")
hdNonVeterans <- sum(nonVeteranHDClean$heartdisease == "Yes")

successes <- c(hdVeterans, hdNonVeterans)
n_obs <- c(nrow(veteranHDClean), nrow(nonVeteranHDClean))

hd_prop_test <- prop.test(successes, n_obs)
hd_prop_test
```

Cancer: p-value $< 2.2 \times 10^{-16}$
```{r}
#| echo: false

veteranCancerClean <- veteran %>% filter(!is.na(cancer))
nonVeteranCancerClean <- nonVeteran %>% filter(!is.na(cancer))

cancerVeterans <- sum(veteranCancerClean$cancer == "Yes")
cancerNonVeterans <- sum(nonVeteranCancerClean$cancer == "Yes")

successes <- c(cancerVeterans, cancerNonVeterans)
n_obs <- c(nrow(veteranCancerClean), nrow(nonVeteranCancerClean))

cancer_prop_test <- prop.test(successes, n_obs)
cancer_prop_test
```

Diabetes: p-value $< 2.2 \times 10^{-16}$
```{r}
#| echo: false

veteranDiabetesClean <- veteran %>% filter(!is.na(diabetes))
nonVeteranDiabetesClean <- nonVeteran %>% filter(!is.na(diabetes))

diabetesVeterans <- sum(veteranDiabetesClean$diabetes == "Yes")
diabetesNonVeterans <- sum(nonVeteranDiabetesClean$diabetes == "Yes")

successes <- c(diabetesVeterans, diabetesNonVeterans)
n_obs <- c(nrow(veteranDiabetesClean), nrow(nonVeteranDiabetesClean))

diabetes_prop_test <- prop.test(successes, n_obs)
diabetes_prop_test
```

COPD: p-value $< 2.2 \times 10^{-16}$
```{r}
#| echo: false

veteranCOPDClean <- veteran %>% filter(!is.na(copd))
nonVeteranCOPDClean <- nonVeteran %>% filter(!is.na(copd))

copdVeterans <- sum(veteranCOPDClean$copd == "Yes")
copdNonVeterans <- sum(nonVeteranCOPDClean$copd == "Yes")

successes <- c(copdVeterans, copdNonVeterans)
n_obs <- c(nrow(veteranCOPDClean), nrow(nonVeteranCOPDClean))

copd_prop_test <- prop.test(successes, n_obs)
copd_prop_test
```

daysphyshlthnotgood: p-value $< 2.2 \times 10^{-16}$
```{r}
#| echo: false

veteranDaysphyshlthnotgoodClean <- veteran %>% filter(!is.na(daysphyshlthnotgood))
nonVeteranDaysphyshlthnotgoodClean <- nonVeteran %>% filter(!is.na(daysphyshlthnotgood))

daysphyshlthnotgoodVeterans <- mean(veteranDaysphyshlthnotgoodClean$daysphyshlthnotgood)
daysphyshlthnotgoodNonVeterans <- mean(nonVeteranDaysphyshlthnotgoodClean$daysphyshlthnotgood)

daysphyshlthnotgood_t_test <- t.test(veteranDaysphyshlthnotgoodClean$daysphyshlthnotgood,
                                     nonVeteranDaysphyshlthnotgoodClean$daysphyshlthnotgood,
                                     alternative="two.sided",
                                     mu=0, paired=FALSE, var.equal = FALSE,
                                     conf.level=0.95)
daysphyshlthnotgood_t_test
```

Check Assumptions (equal variances):
```{r}
#| echo: false
#| 
hist(veteranDaysphyshlthnotgoodClean$daysphyshlthnotgood,
     main = "Days Physical Health Not Good (Veteran)",
     xlab = "Number of days in past month physical health was not good")
hist(nonVeteranDaysphyshlthnotgoodClean$daysphyshlthnotgood,
     main = "Days Physical Health Not Good (Non-Veteran)",
     xlab = "Number of days in past month physical health was not good")
```


```{r}
#| echo: false
# Summary of physical health tests

factors <- c("Arthritis", "Heart Disease", "Cancer", "Diabetes", "COPD", "Days Physical Health Not Good")
prop_vet <- c(0.4356422, 0.1244373, 0.1789827, 0.1962027, 0.11969946, 4.972634)
prop_nonVet <- c(0.3455287, 0.0543102, 0.1105903, 0.1340290, 0.07724301, 4.338173)
p_values <- rep("< 2.2e-16", 6)

physical_tests <- data.frame(
  Factors = factors,
  Veteran = prop_vet,
  NonVeteran = prop_nonVet,
  pvalue = p_values
)

physical_tests
```

```{r}
#| echo: false
# generated with help from gemini

# Your vectors
labels_a <- c("Arthritis", "Heart Disease", "Cancer", "Diabetes", "COPD")
prop_vet <- c(0.4356422, 0.1244373, 0.1789827, 0.1962027, 0.11969946)
prop_nonVet <- c(0.3455287, 0.0543102, 0.1105903, 0.1340290, 0.07724301)

# Combine into a matrix
data_matrix_a <- rbind(prop_vet, prop_nonVet)

# Assign column names to the matrix
colnames(data_matrix_a) <- labels_a

# Your vectors
labels_b <- c("daysphyshlthnotgood")
mean_vet <- c(4.972634)
mean_nonVet <- c(4.338173)

# Combine into a matrix
data_matrix_b <- rbind(mean_vet, mean_nonVet)

# Assign column names to the matrix
colnames(data_matrix_b) <- labels_b

# Create the grouped barplot
# Adjust margins so the long label fits (bottom, left, top, right)
par(mar = c(12, 4, 4, 2) + 0.1)
par(mfrow=c(1,2))

barplot(data_matrix_a, 
        beside = TRUE,                   # This places the bars side-by-side
        col = c("skyblue", "salmon"),    # Colors for the two groups
        legend.text = c("Veteran", "Non-Veteran"),
        args.legend = list(x = "topright", bty = "n"),
        main = "Health Indicators: ",
        ylab = "Value",
        las = 1,                         # Rotates x-axis labels to be vertical
        cex.names = 0.8)                 # Shrinks the label text slightly

barplot(data_matrix_b, 
        beside = TRUE,                   # This places the bars side-by-side
        col = c("skyblue", "salmon"),    # Colors for the two groups
        legend.text = c("Veteran", "Non-Veteran"),
        args.legend = list(x = "topright", bty = "n"),
        main = "Veterans vs Non-Veterans",
        ylab = "Value",
        las = 1,                         # Rotates x-axis labels to be vertical
        cex.names = 0.8)                 # Shrinks the label text slightly
```

Comparisons of mental health between veterans and non-veterans:

Depression: p-value $< 2.2 \times 10^{-16}$
```{r}
#| echo: false

veteranDepressClean <- veteran %>% filter(!is.na(depress))
nonVeteranDepressClean <- nonVeteran %>% filter(!is.na(depress))

depressVeterans <- sum(veteranDepressClean$depress == "Yes")
depressNonVeterans <- sum(nonVeteranDepressClean$depress == "Yes")

successes <- c(depressVeterans, depressNonVeterans)
n_obs <- c(nrow(veteranDepressClean), nrow(nonVeteranDepressClean))

depress_prop_test <- prop.test(successes, n_obs)
depress_prop_test
```

daysmenthlthnotgood: p-value $< 2.2 \times 10^{-16}$
```{r}
#| echo: false

veteranDaysmenthlthnotgoodClean <- veteran %>% filter(!is.na(daysmenthlthnotgood))
nonVeteranDaysmenthlthnotgoodClean <- nonVeteran %>% filter(!is.na(daysmenthlthnotgood))

daysmenthlthnotgoodVeterans <- mean(veteranDaysmenthlthnotgoodClean$daysmenthlthnotgood)
daysmenthlthnotgoodNonVeterans <- mean(nonVeteranDaysmenthlthnotgoodClean$daysmenthlthnotgood)

daysmenthlthnotgood_t_test <- t.test(veteranDaysmenthlthnotgoodClean$daysmenthlthnotgood,
                                     nonVeteranDaysmenthlthnotgoodClean$daysmenthlthnotgood,
                                     alternative="two.sided",
                                     mu=0, paired=FALSE, var.equal = TRUE,
                                     conf.level=0.95)
daysmenthlthnotgood_t_test
```

Check Assumptions (equal variances):
```{r}
#| echo: false
#| 
hist(veteranDaysmenthlthnotgoodClean$daysmenthlthnotgood,
     main = "Days Mental Health Not Good (Veteran)",
     xlab = "Number of days in past month mental health was not good")
hist(nonVeteranDaysmenthlthnotgoodClean$daysmenthlthnotgood,
     main = "Days Mental Health Not Good (Non-Veteran)",
     xlab = "Number of days in past month mental health was not good")
```

```{r}
#| echo: false
# Summary of Mental Health Tests

factors <- c("Depression", "Days Physical Health Not Good")
prop_vet <- c(0.1757344, 3.517690)
prop_nonVet <- c(0.2166535, 4.462462)
p_values <- rep("< 2.2e-16", 2)

mental_tests <- data.frame(
  Factors = factors,
  Veteran = prop_vet,
  NonVeteran = prop_nonVet,
  pvalue = p_values
)

mental_tests
```

```{r}
#| echo: false
# generated with help from gemini

# Your vectors
labels_a <- c("Depression")
prop_vet <- c(0.1757344)
prop_nonVet <- c(0.2166535)

# Combine into a matrix
data_matrix_a <- rbind(prop_vet, prop_nonVet)

# Assign column names to the matrix
colnames(data_matrix_a) <- labels_a

# Your vectors
labels_b <- c("daysmenthlthnotgood")
mean_vet <- c(3.517690)
mean_nonVet <- c(4.462462)

# Combine into a matrix
data_matrix_b <- rbind(mean_vet, mean_nonVet)

# Assign column names to the matrix
colnames(data_matrix_b) <- labels_b

# Create the grouped barplot
# Adjust margins so the long label fits (bottom, left, top, right)
par(mar = c(12, 4, 4, 2) + 0.1)
par(mfrow=c(1,2))

barplot(data_matrix_a, 
        beside = TRUE,                   # This places the bars side-by-side
        col = c("skyblue", "salmon"),    # Colors for the two groups
        legend.text = c("Veteran", "Non-Veteran"),
        args.legend = list(x = "topright", bty = "n"),
        main = "Health Indicators: ",
        ylab = "Value",
        las = 1,                         # Rotates x-axis labels to be vertical
        cex.names = 0.8)                 # Shrinks the label text slightly

barplot(data_matrix_b, 
        beside = TRUE,                   # This places the bars side-by-side
        col = c("skyblue", "salmon"),    # Colors for the two groups
        legend.text = c("Veteran", "Non-Veteran"),
        args.legend = list(x = "topright", bty = "n"),
        main = "Veterans vs Non-Veterans",
        ylab = "Value",
        las = 1,                         # Rotates x-axis labels to be vertical
        cex.names = 0.8)                 # Shrinks the label text slightly
```

What is the difference in the probability of having a specific health factor being a veteran vs. non-veteran?

```{r}
#| echo: false

arthritisClean <- brfss %>% filter(!is.na(arthritis))

arthritisClean$veteran <- as.factor(arthritisClean$veteran)

arthritisClean$arthritis <- as.factor(arthritisClean$arthritis)

arthritisMod <- glm(formula = arthritis ~ veteran, data = arthritisClean, family = "binomial")

summary(arthritisMod)
```

```{r}
p_vet_arthritis <- plogis(-0.63875 + 0.37988 * 1)
p_nonVet_arthritis <- plogis(-0.63875 + 0.37988 * 0)

cat("P(Arthritis | Veteran) = ", p_vet_arthritis)
cat('\n')
cat("P(Arthritis | Non-Veteran) = ", p_nonVet_arthritis)
```

```{r}
#| echo: false

hdClean <- brfss %>% filter(!is.na(heartdisease))

hdClean$veteran <- as.factor(hdClean$veteran)

hdClean$heartdisease <- as.factor(hdClean$heartdisease)

hdMod <- glm(formula = heartdisease ~ veteran, data = hdClean, family = "binomial")

summary(hdMod)
```

```{r}
p_vet_hd <- plogis(-2.857202 + 0.906138 * 1)
p_nonVet_hd <- plogis(-2.857202 + 0.906138 * 0)

cat("P(Heart Disease | Veteran) = ", p_vet_hd)
cat('\n')
cat("P(Heart Disease | Non-Veteran) = ", p_nonVet_hd)
```

```{r}
#| echo: false

cancerClean <- brfss %>% filter(!is.na(cancer))

cancerClean$veteran <- as.factor(cancerClean$veteran)

cancerClean$cancer <- as.factor(cancerClean$cancer)

cancerMod <- glm(formula = cancer ~ veteran, data = cancerClean, family = "binomial")

summary(cancerMod)
```

```{r}
p_vet_cancer <- plogis(-2.084725 + 0.56147 * 1)
p_nonVet_cancer <- plogis(-2.084725 + 0.56147 * 0)

cat("P(Cancer | Veteran) = ", p_vet_cancer)
cat('\n')
cat("P(Cancer | Non-Veteran) = ", p_nonVet_cancer)
```

```{r}
#| echo: false

diabetesClean <- brfss %>% filter(!is.na(diabetes))

diabetesClean$veteran <- as.factor(diabetesClean$veteran)

diabetesClean$diabetes <- as.factor(diabetesClean$diabetes)

diabetesMod <- glm(formula = diabetes ~ veteran, data = diabetesClean, family = "binomial")

summary(diabetesMod)
```

```{r}
p_vet_diabetes <- plogis(-1.602633 + 0.370480 * 1)
p_nonVet_diabetes <- plogis(-1.602633 + 0.370480 * 0)

cat("P(Diabetes | Veteran) = ", p_vet_diabetes)
cat('\n')
cat("P(Diabetes | Non-Veteran) = ", p_nonVet_diabetes)
```

```{r}
#| echo: false

copdClean <- brfss %>% filter(!is.na(copd))

copdClean$veteran <- as.factor(copdClean$veteran)

copdClean$diabetes <- as.factor(copdClean$copd)

copdMod <- glm(formula = copd ~ veteran, data = copdClean, family = "binomial")

summary(copdMod)
```

```{r}
p_vet_copd <- plogis(-2.480409 + 0.485130 * 1)
p_nonVet_copd <- plogis(-2.480409 + 0.485130 * 0)

cat("P(COPD | Veteran) = ", p_vet_copd)
cat('\n')
cat("P(COPD | Non-Veteran) = ", p_nonVet_copd)
```

```{r}
#| echo: false

depressClean <- brfss %>% filter(!is.na(depress))

depressClean$veteran <- as.factor(depressClean$veteran)

depressClean$depress <- as.factor(depressClean$depress)

depressMod <- glm(formula = depress ~ veteran, data = depressClean, family = "binomial")

summary(depressMod)
```

```{r}
p_vet_depress <- plogis(-1.285276 - 0.260243 * 1)
p_nonVet_depress <- plogis(-1.285276 - 0.260243 * 0)

cat("P(Depressive Disorder Diagnosis | Veteran) = ", p_vet_depress)
cat('\n')
cat("P(Depressive Disorder Diagnosis | Non-Veteran) = ", p_nonVet_depress)
```


Elsa Section

What are some factors that influence physical health among veterans

```{r}
library(sjPlot)
library(sjlabelled)
library(sjmisc)
library(ggplot2)
brfss = read.csv("brfss2022_m62.csv", stringsAsFactors=TRUE)

head(brfss)
```

```{r Physical Health: Heart Disease}
#| echo: false
veteran <- subset(brfss, veteran == "Yes")


# Code below is Gemini generated: Prompt was that given these age groups, merge them so that we have three columns of young adult, middle age, and seniors

# 1. Create the new categorization logic
# We assign the categories based on the modified strings
veteran$age_category <- ifelse(veteran$age %in% c("25 to 29", "30 to 34", "35 to 39"), "young adult",
                        ifelse(veteran$age %in% c("40 to 44", "45 to 49", "50 to 54", "55 to 59", "60 to 64"), "middle-aged",
                        ifelse(veteran$age %in% c("65 to 69", "70 to 74", "75 to 79", "80 or older"), "senior", NA)))

# 2. Remove the old granular column and rename the new one to 'age'
veteran$age <- NULL
names(veteran)[names(veteran) == "age_category"] <- "age"

```

```{r}
# Two lines below generated by Gemini, prompt was that given these columns, drop any rows where any of these columns are NA
# Define the variables of interest
vars <- c("heartdisease", "sex", "bmi", "cancer", "copd", 
          "daysphyshlthnotgood", "anyexer", "race", 
          "age", "bingedrinker", "weeklydrinks")

# Subset rows where all specified columns are non-NA
veteran_physical <- veteran[complete.cases(veteran[, vars]), ]


model_heartdiease <- glm(formula = heartdisease ~ sex + bmi + cancer + copd +
                 daysphyshlthnotgood + anyexer + race +
                 age + bingedrinker + weeklydrinks , data=veteran_physical, family=binomial)

#The line below are Gemini generated: Prompt was that given a glm model, create automatic stepwise param selection
# Perform stepwise selection using AIC
# direction = "both" checks for variables to add AND remove at each step
final_model_heartdisease <- step(model_heartdiease, direction = "both")




```
```{r}
# This line is Gemini Generated: Prompt was that given a glm model, create automatic stepwise param selection
# Perform stepwise selection using AIC
# View the final selected model
summary(final_model_heartdisease)
```

```{r}
theme_set(theme_sjplot())
tab_model(final_model_heartdisease, show.stat = FALSE)
```


```{r Physical Health: Cancer}
#| echo: false

model_cancer <- glm(formula = cancer ~ sex + bmi  + copd +
                 daysphyshlthnotgood + anyexer + race +
                 age + bingedrinker + weeklydrinks , data=veteran_physical, family=binomial)

#The two lines below are Gemini generated: Prompt was that given a glm model, create automatic stepwise param selection
# Perform stepwise selection using AIC
# direction = "both" checks for variables to add AND remove at each step
final_model_cancer <- step(model_cancer)

# View the final selected model
summary(final_model_cancer)
tab_model(final_model_cancer, show.stat = FALSE)
```


```{r Physical Health: Diabetes}
#| echo: false

model_diabetes <- glm(formula = diabetes ~ sex + bmi + cancer + copd +
                 daysphyshlthnotgood + anyexer + race +
                 age + bingedrinker + weeklydrinks , data=veteran_physical, family=binomial)

#The two lines below are Gemini generated: Prompt was that given a glm model, create automatic stepwise param selection
# Perform stepwise selection using AIC
# direction = "both" checks for variables to add AND remove at each step
final_model_diabetes <- step(model_diabetes)

# View the final selected model
summary(final_model_diabetes)
tab_model(final_model_diabetes, show.stat = FALSE)
```
What are some factors that influence mental health among veterans

```{r Mental Health: Depression}
#| echo: false

# Two lines below generated by Gemini, prompt was that given these columns, drop any rows where any of these columns are NA
# Define the variables of interest

vars_mental <- c("daysmenthlthnotgood", "medcost", "income", "race", "age", "bingedrinker", "weeklydrinks", "sex", "depress")

# Subset rows where all specified columns are non-NA
veteran_mental <- veteran[complete.cases(veteran[, vars_mental]), ]


model_depression <- glm(formula = depress ~ sex + medcost + income + race +
                 age + bingedrinker + weeklydrinks + daysmenthlthnotgood , data=veteran_mental, family=binomial)

#The two lines below are Gemini generated: Prompt was that given a glm model, create automatic stepwise param selection
# Perform stepwise selection using AIC
# direction = "both" checks for variables to add AND remove at each step
final_model_depression <- step(model_depression, direction = "both")

# View the final selected model
summary(final_model_depression)
tab_model(final_model_depression, show.stat = FALSE)
```


```{r Mental Health: Daysmenthlthnotgood}
#| echo: false

# Two lines below generated by Gemini, prompt was that given these columns, drop any rows where any of these columns are NA
# Define the variables of interest
model_daysmenthlthnotgood<- lm(formula = daysmenthlthnotgood ~ sex + medcost + income + race +
                 age + bingedrinker + weeklydrinks + depress , data=veteran_mental)

#The two lines below are Gemini generated: Prompt was that given a glm model, create automatic stepwise param selection
# Perform stepwise selection using AIC
# direction = "both" checks for variables to add AND remove at each step
final_model_daysmenthlthnotgood <- step(model_daysmenthlthnotgood, direction = "both")

# View the final selected model
summary(final_model_daysmenthlthnotgood)
tab_model(final_model_daysmenthlthnotgoodx, show.stat = FALSE)
```